% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
% fallback for those not using the hyperref driver hyperxmp:
\makeatletter
\@ifundefined{xmpquote}{\newcommand{\xmpquote}[1]{#1}}{}
\makeatother
\hypersetup{
  pdftitle={xlq: A Safe-Write Boundary for LLM Agents Editing Spreadsheets},
  pdfauthor={the xlq authors},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{xlq: A Safe-Write Boundary for LLM Agents Editing Spreadsheets}
\author{the xlq authors}
\date{2026-07-04}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\section{xlq: A Safe-Write Boundary for LLM Agents Editing
Spreadsheets}\label{xlq-a-safe-write-boundary-for-llm-agents-editing-spreadsheets}

\emph{An Enforced Fidelity Property, an Oracle-Gated Write Path, and an
Adversarial Confinement Trial}

\emph{Authors omitted for review.}

\subsection{Abstract}\label{abstract}

Given the same one-cell edit on the same workbook, a status-quo LLM
agent using the standard Python library produced a corrupt,
non-reloadable file --- two charts and a pivot table stripped of their
relationships, 13 of 50 package parts dropped, only 1 of 50 left
byte-identical --- and \emph{reported success}, having read the raw XML
and confirmed its cell had landed; an agent confined to the boundary
presented here made the identical edit with charts and pivot
byte-identical, 48 of 50 parts untouched, the file reloadable, and a
signed receipt. This is the corruption of Anthropic\textquotesingle s
own shipped spreadsheet skill (issue \#22044), reproduced live under a
real agent, and prevented. The boundary is xlq, and its core is a
\emph{surgical} write primitive with a fidelity property that holds by
construction and is enforced on every write: after an edit, every OOXML
part that does not contain a changed cell is byte-identical to the
input, because the writer copies those parts verbatim and re-serializes
only the sheet parts an operation touches --- and because every apply
byte-diffs the un-edited parts before it commits and aborts on any
change --- a property that openpyxl, LibreOffice, and even the
engine\textquotesingle s own whole-file writer all fail, each rewriting
the entire container. The write is wrapped in a transactional envelope:
a \texttt{base\_hash} precondition under an advisory lock, a
\texttt{-\/-dry-run} that predicts affected cells and new formula
errors, a proof-carrying commit that re-loads its own output, verifies
every predicted cell landed, and confirms every non-edited part survived
byte-identical before it commits, a write-reliability gate that refuses
to persist any engine-computed cache whose formula uses a function our
own differential oracle found divergent from Excel, an atomic revision
swap, and an append-only hash-chained receipt journal. We define the
enforcement claim precisely --- an agent whose harness confines its
workbook writes to xlq cannot reach the \#22044 failure mode, while an
agent handed a raw shell can, which is the harness\textquotesingle s
responsibility and not xlq\textquotesingle s --- matching how the
2025--26 agent-enforcement literature scopes its own boundaries.
Underneath, the boundary reads a workbook through a privacy-safe
structural census (999 B versus a 7.24 MB full-cell dump on a
\textasciitilde100k-formula file), reports its own per-artifact
evaluability as three honest numbers (522 of 522 catalog functions
recognized, 505 locally evaluable, 17 policy-limited), and rests on a
formula engine validated by a documentation-arbitered differential
oracle over 1,659 cases; an independent financial cross-check confirms
two of the oracle\textquotesingle s Treasury-bill verdicts and
\emph{downgrades} a third, which we report as a strength of the method
rather than hide. All artifacts are reproducible.

\subsection{1. Introduction}\label{1-introduction}

Anthropic\textquotesingle s own Claude Code shipped a spreadsheet skill
that quietly destroyed the files it edited (issue \#22044): asked to
change a few cells in a financial model, the agent completed the
requested edit and reported success, but the workbook it saved had lost
its charts, pivot tables, and macros --- stripped by openpyxl, the
standard Python library the skill used to write the file back to disk.
The loss was invisible at exactly the moment it mattered: the
agent\textquotesingle s own report, the diff a reviewer would skim, and
the cells the task touched all looked correct. When an LLM agent edits a
spreadsheet, its decisions do not sit in a separate log --- they become
the cells the file\textquotesingle s owner is accountable for {[}Panko
2008; Hermans and Murphy-Hill 2015{]}. The agent had no safe way to
write the file. Its only primitive was to load the whole workbook into a
library that models a subset of the format, mutate a cell, and
re-serialize the entire container --- discarding everything the library
does not model as a side effect of saving.

An earlier version of this work (§9) diagnosed that failure and built a
read-only inspector that could \emph{observe} the damage but not prevent
it: a boundary that cannot write cannot corrupt, but it also cannot let
an agent do the job it was asked to do. That is not an enforcement
boundary; it is an abstention. This paper builds the missing primitive.
The contribution is a \emph{surgical transactional write} --- a way for
an agent to change a cell that rewrites only the sheet parts an edit
actually touches and copies every other part of the file byte-for-byte
--- together with the transactional envelope (precondition, dry-run,
atomic revision, receipt) that makes each edit previewable,
attributable, and reversible, and the threat model that says precisely
whom it protects against and whom it does not. The property that makes
it a boundary rather than a heuristic holds by construction and is
checked at write time, and is the spine of the paper:

\begin{quote}
\textbf{Fidelity property.} After \texttt{xlq\ apply}, every OOXML part
that does not contain a cell changed by the patch is byte-identical to
the corresponding input part.
\end{quote}

Charts, pivot caches, VBA projects, drawings, styles, theme, and
external links are preserved not because the writer is careful with them
but because it never touches them --- it streams them from the input zip
to the output zip unchanged. This is exactly the property openpyxl,
LibreOffice, and IronCalc\textquotesingle s own \texttt{save\_to\_xlsx}
all fail, because each deserializes the model it understands and
re-serializes the whole package (§3). It directly closes the \#22044
harm class.

The central evidence is interventional, not static --- a
demonstration-scale case study, not a rate. We ran a real LLM agent on
the same edit task, on the same workbook, changing only the tool it was
given to write the file. Through the status-quo openpyxl path the agent
produced a workbook that will not reopen --- charts and pivot stripped
of their relationships, 13 of 50 parts dropped, 1 of 50 byte-identical
--- and reported success anyway, having read the raw sheet XML and seen
its target cell written; through xlq the agent made the identical edit
with the charts and pivot byte-identical, 48 of 50 parts untouched, the
file reloadable, and a hash-chained receipt on disk (§4). The agent,
prompt, input file, and environment were held fixed; only the write tool
varied, so in this pair of trials the difference in outcome is
attributable to the tool boundary. In the trial, an agent confined to
xlq could not commit the \#22044 corruption \emph{even when it tried the
identical edit} --- a difference that is mechanical, not a matter of the
agent\textquotesingle s care.

The rest of the boundary supports that write path and is largely carried
over from the read-only system, now in service of a tool that writes.
The workbook is read through a privacy-safe structural census, so an
agent sees a file\textquotesingle s structure without its contents at a
fraction of the token cost of dumping cells (999 bytes versus a
7,239,303-byte full-cell dump on a \textasciitilde100k-formula workbook,
a 7,246.5× reduction). The runtime reports its own per-file evaluability
as three honest numbers rather than one --- 522 of 522 catalog functions
recognized, 505 locally evaluable, 17 returning the exact error literal
desktop Excel returns --- and the dry-run degrades that flag when a
volatile or unsupported function lies in an edit\textquotesingle s
dependency graph, so a write whose predicted values cannot be trusted is
refused rather than shipped. The formula engine that powers both the
dry-run prediction and the differential comparisons is validated by a
documentation-arbitered differential oracle over 1,659 hand-authored
cases; a new independent financial cross-check (§7) confirms two of its
Treasury-bill verdicts and downgrades a third, which we report openly
because the cross-check caught one of our own calls.

This paper makes five contributions. \textbf{First}, the surgical write
primitive and its fidelity property --- enforced and \emph{checked per
apply}, each write re-loading its own output as an on-disk
\texttt{.xlsx} to confirm every predicted cell landed \emph{and}
byte-diffing every non-edited OOXML part against the input, aborting
with \texttt{fidelity\_violation} if any un-edited part changed, so the
property is checked at write time rather than only asserted by
construction --- with per-file preservation results on four real
workbooks carrying charts, pivots, and VBA (§3). \textbf{Second}, an
interventional case study --- n = 2 tasks, one model, one trajectory per
arm --- that reproduces the \#22044 corruption in a single observed
instance under a status-quo agent and shows an xlq-confined agent
mechanically unable to commit it, including the false-success failure
where the agent believed it had succeeded (§4); this is
demonstration-scale evidence that the harm and its prevention exist, not
a rate or a generality claim. \textbf{Third}, an explicit threat model
whose non-bypassability we tested adversarially in a single
adversarial-agent trial rather than only argued: an adversarial agent
inside a confined harness could not forge a valid-but-wrong workbook by
any route but xlq, surfaced a blunt-destruction hole, and had it closed
by a read-only authoritative store and a receipt-gated broker --- the
boundary confines an agent routed through it, not one handed raw shell
(§5). One trial that found and closed a real hole is evidence, not a
proof. \textbf{Fourth}, the format-aware reading layer --- the
privacy-safe census and the three-number coverage taxonomy --- that lets
a write be previewed and its trustworthiness declared per file (§6).
\textbf{Fifth}, a documentation-arbitered differential oracle for a
domain with no mechanized specification, now cross-checked against an
independent financial reference that retires the circularity objection
for two verdicts and honestly downgrades a third --- and whose confusion
matrix is now \emph{load-bearing for write safety}, the write refusing
to persist any engine-computed cache whose formula uses an
oracle-divergent function (§7), packaged with the rest as AXLE-bench,
whose fidelity axis measures the dimension current agent benchmarks omit
(§8). We are explicit about what we did not invent: receipts,
transactional isolation, and differential testing are established
{[}Notarized Agents 2026; Fault-Tolerant Sandboxing 2025; Yang et al.
2011{]}; the contribution is folding the artifact\textquotesingle s
\emph{format fidelity} into an agent write boundary --- surgically, with
an enforced-and-checked fidelity property, and demonstrated under a real
agent --- for a legacy artifact class where no safe write primitive
previously existed.

\textbf{Scope, stated honestly.} This is a tool-and-systems contribution
at demonstration scale, and we mark its edges rather than let strong
verbs outrun the evidence. What is built and tested: (a) a safe-write
primitive for \emph{in-place cell and formula edits}, whose fidelity
property is enforced and checked on every apply --- the writer copies
un-edited parts verbatim by construction, and before any commit the
apply byte-diffs every non-edited OOXML part against the input and
aborts with \texttt{fidelity\_violation} on any change; (b) an
interventional case study (n = 2 tasks, one model, one trajectory per
arm) that reproduces the \#22044 corruption in a single observed
instance and shows an xlq-confined agent mechanically unable to repeat
it; (c) an adversarially-tested confinement mechanism, exercised in one
adversarial-agent trial that found and closed a single real hole; and
(d) concrete formula bugs surfaced in two independent engines. What is
\emph{not} here: the primitive covers only value and formula edits ---
\emph{structural edits (row/column insert and delete, sheet add/remove,
range moves), the operations most likely to corrupt a workbook, are
unbuilt and are named as the key open problem}; and the agent evidence
is demonstration-scale --- an existence demonstration that the harm and
its prevention are real and reproducible under a real agent, never a
corruption rate or a claim of generality. We write "enforced" and
"checked" where the code enforces and checks, "case study" and "trial"
where n is small, and we claim no formal proof anywhere.

\subsection{2. The Fidelity Gap}\label{2-the-fidelity-gap}

When an LLM agent edits a workbook, the edit is not written to a log
beside the artifact --- it \emph{becomes} the artifact. The cells the
agent produces are the cells the file\textquotesingle s owner will sign
off on, submit to an auditor, and be held accountable for {[}Panko 2008;
Hermans and Murphy-Hill 2015{]}. This promotes a property that
task-success evaluation never names into a first-class safety concern:
\emph{fidelity} --- whether the parts of the file the agent did not
intend to touch survive the operation unchanged. This section
establishes, from the failure of the field\textquotesingle s own
precedents, that fidelity is at once routinely disturbed by the
substrate agents use and systematically unmeasured by the benchmarks
they are scored against. It also fixes the measurement discipline the
rest of the paper rests on: byte-identity and cached values are
untrustworthy across every whole-file substrate, so any correct
integrity check must be semantic and recompute-aware (claim C1) --- and,
as §3 shows, the only way to \emph{keep} byte-identity as a signal is to
stop rewriting the whole file.

\textbf{Method.} We took a five-workbook corpus and re-saved each file
through three substrates: openpyxl 3.1.5, LibreOffice 24.8.7.2
(\texttt{-\/-convert-to\ xlsx}, headless), and the vendored IronCalc
engine (master \texttt{e50ccea8}). Each output was compared to its input
at two levels: an OOXML part census (part names, uncompressed sizes, and
content CRC32s from the zip central directory) and a cell-level
\emph{recompute-aware} semantic diff (§6), which recomputes every
formula in the vendored engine and reports a distinct
\texttt{cached\_value} change-kind when a cell\textquotesingle s stored
cache differs from its input while its live recomputed value does not.
One provenance caveat governs the IronCalc row: the fixtures were
authored by IronCalc itself, so that row is a \emph{best case} for the
engine\textquotesingle s self-round-trip; a companion run shows it
rewrites foreign files as freely as the others. IronCalc is not the hero
of this table; it is the row that shows what "no change" looks like as a
baseline --- and the row §3 exists to improve on.

\textbf{Byte-identity is destroyed everywhere.} Re-saving rewrites
essentially the whole package regardless of tool. LibreOffice rewrote
\emph{every} part of every fixture --- not one part survived
byte-identical. openpyxl rewrote all but the static theme part, and
dropped \texttt{xl/sharedStrings.xml} --- but this is normalization, not
loss: openpyxl re-inlines the strings into the sheet XML (verified: 170
shared-reference cells became 170 inline-string cells with every text
value present), so information is conserved while byte layout is not.
The consequence is decisive: because \textasciitilde100\% of
content-bearing OOXML parts are rewritten by both mainstream substrates,
a byte-level or CRC-level diff reports "everything changed" on a re-save
that changed no user-visible value, and is therefore useless as an
integrity oracle. A check that can separate a benign re-serialization
from a value-altering edit cannot operate on bytes --- \emph{unless a
tool refuses to rewrite the parts it did not edit}, which is precisely
what §3 does.

\textbf{Three tiers, measured separately.} Once byte-diff is discarded,
the recompute-aware diff resolves the transformations into three
distinct tiers, which must not be collapsed into a single word like
"damage" (Table 1):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lllll@{}}
\toprule\noalign{}
Substrate (version) & OOXML parts left byte-identical & Cached formula
values affected & Character & Tier \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
openpyxl 3.1.5 & theme only & \textbf{101,961} blanked to
\texttt{\textless{}v/\textgreater{}} & recoverable by recomputation
(formulas intact); invisible to any cache-trusting consumer & T2.5 \\
LibreOffice 24.8.7.2 & none & \textbf{90,448} rewritten at 15 vs. 17
significant digits & below the precision of an IEEE-754 double;
recomputed on open & T3 \\
IronCalc master \texttt{e50ccea8} & all (self-authored) & 0 & best-case
baseline; rewrites foreign files similarly & --- \\
\end{longtable}
}

\emph{Table 1. Re-save fidelity across substrates, five-workbook corpus
(results.json, D).}

\emph{T2 (semantics-preserving normalization).}
openpyxl\textquotesingle s shared-string re-inlining is lossless; its
only real effect is to defeat byte-provenance, already covered above.

\emph{T2.5 (recoverable, but breaks cache-trusting consumers).} openpyxl
blanks the cached result of \textbf{every} formula it round-trips ---
\textbf{101,961} \texttt{\textless{}v/\textgreater{}} cells across the
corpus (results.json, D; 442 in \texttt{branch-consolidation.xlsx},
99,800 concentrated in the \textasciitilde100k-formula
\texttt{perf-large.xlsx}, present in all five files). Because the
formula text survives, the values are recoverable by recomputation; but
any downstream consumer that reads the cache rather than recomputing ---
a viewer, an importer, a second agent, a \texttt{git\ diff}-style
comparator --- now sees empty cells where numbers were.

\emph{T3 (cosmetic sub-precision drift).} LibreOffice writes formula
caches at 15 rather than 17 significant digits --- \textbf{90,448} cells
(results.json, D; almost all in \texttt{perf-large.xlsx}). This is a
legitimate serialization choice, not a defect; the drift is below the
round-trip precision of a double and is recomputed on open. Its only
consequence is, again, that it defeats naive byte- and cache-level
provenance. That both a value-erasing transformation (T2.5) and a benign
precision choice (T3) surface through the \emph{same}
\texttt{cached\_value} change-kind, and are told apart only by
recomputation, is itself the argument for making the diff semantic.

\textbf{T1, the severest tier, needs the write path to measure.} The one
tier the re-save corpus does \emph{not} measure is the most severe: T1,
genuine irreversible loss of charts, pivot tables, or VBA. The synthetic
fixtures used for Table 1 contain none of these features (IronCalc
cannot author them), so Table 1 makes no measured T1 claim and points to
the external evidence that it is real (\#22044). Section 3 closes this
gap directly: it introduces a purpose-built corpus that \emph{does}
carry charts, pivots, and VBA, and measures T1 loss under each tool ---
which is only possible once there is a surgical writer to compare the
whole-file substrates against.

\textbf{The evaluation half of the gap.} None of this is caught by how
spreadsheet agents are scored. SheetCopilot {[}Li et al. 2023{]},
SheetAgent {[}SheetAgent 2025{]}, SpreadsheetBench {[}SpreadsheetBench
2024{]} and its 2026 successor {[}SpreadsheetBench 2026{]}, and Finch
{[}Finch 2025{]} all grade against target cells and are silent on the
untouched remainder; SheetCopilot\textquotesingle s checker skips
unflagged properties \emph{by design}. The single agent benchmark that
scores any side effect at all, OS-Harm {[}OS-Harm 2025{]}, judges harm
categories in the agent\textquotesingle s actions, not fidelity of the
file it leaves behind. An agent could blank all 101,961 caches, report
"done," and score a clean pass on every one of them.

\textbf{Claim C1.} Across every whole-file substrate an agent might use,
re-saving an \texttt{.xlsx} destroys byte-identity and rewrites or
blanks the cached values on which byte- and cache-level integrity checks
depend; distinguishing a benign re-serialization from a value-changing
edit therefore requires a semantic, recompute-aware comparison, and no
existing spreadsheet-agent benchmark measures this dimension. The rest
of the paper is the write primitive that escapes this regime --- by not
rewriting the whole file --- and the evaluation that follows from taking
that claim seriously.

\subsection{3. The Surgical Write}\label{3-the-surgical-write}

The status-quo write path is \emph{whole-file}: load the workbook into a
library, mutate a cell, re-serialize the entire OOXML container.
Everything the library does not model --- chart theming, pivot caches,
VBA, printer settings, the shared-string layout --- is rewritten or
dropped as a side effect of the save, regardless of what the edit was
(§2). The surgical write inverts this. It rewrites only the OOXML parts
that contain an affected cell and copies every other part of the input
zip to the output byte-for-byte. The claim staked here is narrow and, we
argue, new: to our knowledge xlq is the first LLM-agent write boundary
that is aware of its target artifact\textquotesingle s \emph{format
fidelity} --- not merely whether a write is permitted, but whether it
corrupts the artifact. The 2024--26 runtime-enforcement wave established
that a deterministic runtime beats prompt-level guidance {[}Progent
2025; AgentSpec 2025; IsolateGPT 2024{]}, but every system in that wave
is format-blind: it gates a filesystem write without understanding
whether the write preserves the file it guards.

\subsubsection{The fidelity property}\label{the-fidelity-property}

The primitive
\texttt{xlq\ apply\ \textless{}book.xlsx\textgreater{}\ \textless{}patch.json\textgreater{}\ {[}-\/-dry-run{]}}
performs a surgical edit whose defining guarantee is:

\begin{quote}
After \texttt{xlq\ apply}, every OOXML part that does not contain a cell
changed by the patch is \textbf{byte-identical} to the input part.
\end{quote}

The property holds \emph{by construction}, not by care. The writer
\texttt{ooxml::surgical\_write} streams every zip member of the input to
the output unchanged and re-serializes only the sheet parts named in the
edit set; a part it does not touch is copied byte-for-byte, so charts,
pivot caches, and \texttt{vbaProject.bin} cannot be altered by an edit
that does not name their sheet. Three kinds of support back the
property, and it is worth keeping them apart rather than collapsing them
into the word "proof" (docs/FIDELITY.md). It is \emph{asserted by
construction} --- the writer\textquotesingle s control flow copies
un-named parts verbatim; it is \emph{measured on fixtures} --- a
byte-for-byte part diff against the untouched original confirms the
byte-identical-elsewhere half on every fixture below; and, new in this
version, it is \emph{enforced and checked per apply} --- before any
commit, every real apply re-loads its own output as an on-disk
\texttt{.xlsx} and confirms each predicted cell landed with its
predicted value (the edited half), \emph{and} byte-diffs every
non-edited part against the input, aborting the commit with
\texttt{fidelity\_violation} if any part that contains no edited cell
was not preserved byte-identical (the byte-identical-elsewhere half).
This upgrades both halves from "the writer intended it" to a per-write
check: the edited cells verifiably compute, and the un-edited parts are
verifiably untouched, on the file the consumer will open. What the
per-apply check does \emph{not} establish is that a rewritten sheet
\emph{had} to be rewritten (the benchmark\textquotesingle s "rewritten"
set is defined purely as "bytes differ," so it cannot distinguish a
necessary recompute from a spurious reserialization); that residual ---
minimality of the rewrite, not preservation of the rest --- is the only
half still resting on the construction. The one deliberate exception the
check whitelists is \texttt{xl/calcChain.xml}, a derived
formula-ordering cache that a targeted edit makes stale; the writer
drops it (Excel and every engine regenerate it on open), and the fixture
tables below count that drop honestly as "not byte-identical" rather
than hide it.

\subsubsection{The algorithm and the transactional
envelope}\label{the-algorithm-and-the-transactional-envelope}

A single \texttt{apply} runs seven steps
(docs/specs/v02-architecture.md):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Precondition.} \texttt{sha256(file)\ ==\ patch.base\_hash},
  else the operation aborts with \texttt{revision\_mismatch}. The patch
  carries the hash of the file it was computed against, so an edit built
  on a stale view of the workbook is refused, not silently applied.
\item
  \textbf{Lock.} An advisory lock (\texttt{book.xlsx.xlq.lock},
  \texttt{O\_EXCL}) is held across the whole check→write sequence, so a
  concurrent writer gets \texttt{lock\_held} rather than interleaving
  (no time-of-check/time-of-use window).
\item
  \textbf{Predict (also the \texttt{-\/-dry-run} path).} A \emph{copy}
  of the workbook is loaded into IronCalc, the typed operations
  (\texttt{set\_cell}, \texttt{set\_formula}, with an explicit
  JSON-to-Excel type mapping so the number-versus-text and date-serial
  traps are handled, never inferred) are applied, the model is
  evaluated, and the full set of affected cells, their new values, any
  newly introduced formula errors, before/after values for caller-named
  watch cells, and a coverage flag are computed. \texttt{-\/-dry-run}
  returns this prediction and stops, writing nothing. Crucially, the
  coverage flag degrades when a volatile or unsupported function lies in
  the affected dependency graph, and a real apply whose predicted cached
  values are unreliable is \emph{refused} --- the write is
  coverage-gated (§6). A second, sharper gate refuses the write
  (\texttt{coverage\_unreliable},
  \texttt{oracle\_divergent\_functions:\ {[}...{]}}) when the affected
  dependency graph uses a function our own differential oracle (§7)
  caught the engine computing wrong: xlq will not persist an
  engine-computed cache it has documented reason to believe diverges
  from Excel. Both gates fire only on \emph{engine-computed} caches; a
  literal \texttt{set\_cell} writes exactly the value the user typed and
  is never gated, because that value does not depend on the engine at
  all.
\item
  \textbf{Surgical write.} The input \texttt{.xlsx} is opened as a zip.
  Affected cells are mapped to their sheet parts
  (\texttt{xl/worksheets/sheetN.xml}) via \texttt{workbook.xml} and its
  rels. For each affected sheet part, the XML is parsed and only the
  \texttt{\textless{}c\textgreater{}} elements for affected cells are
  edited (setting
  \texttt{\textless{}f\textgreater{}}/\texttt{\textless{}v\textgreater{}}
  and the shared-string or inline value), preserving every other element
  in that sheet --- merged cells, conditional formatting, column widths.
  \texttt{xl/calcChain.xml} is dropped if present. Every other part is
  copied byte-for-byte.
\item
  \textbf{Deterministic repackage.} Fixed zip entry mtimes and a stable
  member ordering make \texttt{result\_hash} reproducible.
\item
  \textbf{Proof-carrying verification and fidelity enforcement.} Before
  anything is committed, two checks run against the repackaged output,
  and either can abort the commit while the authoritative file is still
  untouched (nothing has reached it yet, because the commit at step 7 is
  the only writer). \emph{First}, the output is re-loaded from disk the
  way a downstream consumer would open it --- as an \texttt{.xlsx}, in a
  fresh engine instance --- and every cell the dry-run predicted is
  re-checked against its predicted value; if the output does not re-open
  (\texttt{verification\_failed:\ does\ not\ re-open}) or any predicted
  cell disagrees (\texttt{verification\_failed:\ mismatches{[}...{]}}),
  the commit aborts. \emph{Second}, the apply byte-diffs every part of
  the output against the input and aborts with
  \texttt{fidelity\_violation} if any part that contains no edited cell
  was changed, dropped, or added --- the only whitelisted difference
  being an edited sheet part or the deliberately-dropped
  \texttt{xl/calcChain.xml}. This second check \emph{enforces} the
  fidelity property per write rather than trusting the
  writer\textquotesingle s construction: were the surgical writer ever
  to disturb an un-edited part, the commit would refuse rather than ship
  it. The apply response carries
  \texttt{verified:\ \{reopened,\ cells\_checked,\ all\_landed,\ fidelity\_enforced,\ non\_edited\_parts\_byte\_identical\}},
  and a real apply proceeds only when both are satisfied. Together the
  two checks close the gap between "the writer intended the edit" and
  "the written file verifiably computes the edit and preserves
  everything else": a malformed \texttt{\textless{}c\textgreater{}}, a
  wrong cached value, a corrupt insertion, or a stray touch to a chart
  part is caught here rather than asserted by a \texttt{result\_hash}
  the consumer cannot check.
\item
  \textbf{Atomic apply.} An immutable \texttt{book.rev-N.xlsx} is
  written and fsync\textquotesingle d, then atomically renamed onto
  \texttt{book.xlsx}. A receipt is appended to the hash-chained journal
  \texttt{book.xlsx.xlq.jsonl} ---
  \texttt{\{rev,\ base\_hash,\ result\_hash,\ ops,\ ts,\ actor,\ engine\_version,\ clock,\ seed,\ kind\}}
  --- where each receipt binds the pre-image hash to the post-image
  hash. An edit made outside the tool breaks the chain detectably and is
  recorded as an \texttt{external\_edit} receipt that \emph{adopts} the
  new hash rather than wedging the file.
\end{enumerate}

None of these ingredients is individually new, and the paper claims none
of them (§9). Append-only hash-chained receipts descend from
forward-secure logging {[}Nitro 2025{]}, Certificate Transparency
{[}Certificate Transparency 2022{]}, and W3C PROV {[}Moreau et al.
2013{]}; agent-action receipts from Notarized Agents {[}Notarized Agents
2026{]}; transactional rollback from ACID-snapshot and
compensating-transaction sandboxes {[}Fault-Tolerant Sandboxing 2025;
SAFEFLOW 2025{]}. The journal is a domain instantiation --- a
single-writer local sidecar whose threat model is file custody and
out-of-band edit detection, not receiver attestation --- and its
differentiator over a generic transactional sandbox is that the
dry-run\textquotesingle s prediction is \emph{semantic} (affected cells,
new formula errors, watched magnitudes), not merely atomic. The write
path is implemented and tested: 128 tests (100 unit, plus 5 and 7
integration) exercise the precondition, lock, dry-run, surgical writer,
journal, the proof-carrying verification (its positive path and both
negative paths --- a wrong cached value caught, an unloadable output
caught), and the oracle-divergence write gate.

\subsubsection{E1: per-file fidelity
preservation}\label{e1-per-file-fidelity-preservation}

We take one representative edit per fixture ("set this data cell to that
number" --- the edit an agent actually makes), apply the \emph{same
logical edit three ways}, and diff the resulting OOXML package against
the untouched original at the level of individual zip members
(docs/FIDELITY.md, benchmarks/fidelity.json). The three arms are
\texttt{xlq\ apply} (surgical typed patch), openpyxl
(\texttt{load\_workbook} → set cell → \texttt{save}; for
\texttt{macro.xlsm} this uses the \textbf{default
\texttt{keep\_vba=False}}), and LibreOffice
(\texttt{soffice\ -\/-convert-to} re-save, a proxy --- LibreOffice has
no convenient CLI cell-editor, so its 100\%-rewrite figure is an
\emph{upper bound on re-save churn}, not a like-for-like single-cell
edit). Following the discipline of the experiment, results are reported
\textbf{per file, never summed into a corpus total} --- the interesting
fact is per-workbook (charts survive here, VBA survives there), and
averaging would hide exactly the failures the experiment exists to
surface. The four fixtures are hand-picked to carry the features
whole-file rewriters destroy --- charts, pivot tables and caches, VBA,
cross-sheet formulas --- so E1 is an \emph{existence demonstration} that
the fidelity property holds on real workbooks with those features, per
file, not a population estimate over any workbook distribution.

\textbf{\texttt{pivot-chart.xlsx}} --- 2 charts, 1 pivot table, 2 pivot
caches (50 parts). Edit: \texttt{Sheet1!A2} 222 → 999.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}llllllll@{}}
\toprule\noalign{}
Tool & byte-identical & rewritten & dropped & added & charts & pivot &
reloads (ironcalc) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{xlq} & \textbf{48 / 50} & 1 & 1 & 0 & ✅ byte-identical & ✅
byte-identical & ✅ \\
openpyxl & 1 / 50 & 36 & 13 & 4 & ⚠️ degraded (6 support parts dropped)
& ⚠️ rewritten & ❌ \\
LibreOffice & 0 / 50 & 37 & 13 & 6 & ⚠️ degraded (same 6 support parts
dropped) & ⚠️ rewritten & ✅ \\
\end{longtable}
}

xlq rewrites only the edited sheet and drops the stale
\texttt{calcChain.xml}; both charts, the pivot table, both pivot caches,
shared strings, styles, drawings, comments, and printer settings are
byte-identical. openpyxl drops six chart-support parts (both
\texttt{chart*.xml.rels}, both \texttt{colors*.xml}, both
\texttt{style*.xml}), plus \texttt{sharedStrings.xml}, both comment
parts, both VML drawings, \texttt{printerSettings1.bin}, and
\texttt{calcChain.xml}; the core chart and pivot parts are present but
rewritten, and \textbf{the output does not re-open in the ironcalc
engine}. LibreOffice rewrites all 50 parts and, on this axis, is no
better than openpyxl --- it drops the same six chart-support parts,
inside a package that still re-opens.

\textbf{\texttt{macro.xlsm}} --- VBA macros (11 parts). Edit:
\texttt{Data!B2} 100 → 200.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lllllll@{}}
\toprule\noalign{}
Tool & byte-identical & rewritten & dropped & added & VBA & reloads
(ironcalc) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{xlq} & \textbf{10 / 11} & 1 & 0 & 0 & ✅ \texttt{vbaProject.bin}
byte-identical & ✅ \\
openpyxl & 1 / 11 & 8 & 2 & 0 & ❌ \textbf{dropped} (saved as
\texttt{.xlsx}) & ✅ \\
LibreOffice & 0 / 11 & 11 & 0 & 1 & ✅ present (rewritten) & ✅ \\
\end{longtable}
}

xlq rewrites only the edited sheet; \texttt{vbaProject.bin} is
byte-identical. openpyxl with its default \texttt{keep\_vba=False} drops
\texttt{vbaProject.bin} --- \textbf{all macros gone} --- and
\texttt{sharedStrings.xml}, and can only write \texttt{.xlsx}: the
macro-enabled workbook is silently downgraded to a macro-free one.
LibreOffice keeps the macro but rewrites every part.

\textbf{\texttt{payroll.xlsx}} --- 3 sheets, cross-sheet formulas (13
parts). Edit: \texttt{Rates!B2} 16 → 25 (a rate that feeds the
\texttt{Payroll} sheet).

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lllll@{}}
\toprule\noalign{}
Tool & byte-identical & rewritten & dropped & added \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{xlq} & \textbf{11 / 13} & 2 & 0 & 0 \\
openpyxl & 1 / 13 & 10 & 2 & 0 \\
LibreOffice & 0 / 13 & 12 & 1 & 1 \\
\end{longtable}
}

xlq rewrites \emph{exactly two} worksheet parts --- the edited
\texttt{Rates} sheet and the \texttt{Payroll} sheet, whose formula cells
recompute because they depend on the edited rate; both rewrites carry
changed cell values, and nothing else moves. openpyxl drops
\texttt{sharedStrings.xml} and \texttt{xl/metadata.xml} and rewrites 10
parts.

\textbf{\texttt{claims.xlsx}} --- 2 sheets, \textasciitilde1.3k formulas
(12 parts). Edit: \texttt{Claims!D2} 13335 → 20000.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lllll@{}}
\toprule\noalign{}
Tool & byte-identical & rewritten & dropped & added \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{xlq} & \textbf{11 / 12} & 1 & 0 & 0 \\
openpyxl & 1 / 12 & 9 & 2 & 0 \\
LibreOffice & 0 / 12 & 11 & 1 & 1 \\
\end{longtable}
}

xlq rewrites only the sheet carrying the edit and its downstream
formulas; the \texttt{Limits} sheet, styles, shared strings, and
metadata are byte-identical. openpyxl again drops
\texttt{sharedStrings.xml} and \texttt{metadata.xml}.

\textbf{What the tables establish, stated straight.} The measured
\texttt{feature\_survival} flag records \emph{presence}, not integrity:
a ✅ for openpyxl\textquotesingle s charts means a \texttt{chartN.xml}
is present in the output, not that it is byte-identical ---
openpyxl\textquotesingle s charts are present yet stripped of their
theming and relationship parts, which is why that file will not re-open.
Only xlq\textquotesingle s ✅s are also byte-identical. The reload
failure is reported narrowly and conditionally: ironcalc is
xlq\textquotesingle s own vendored engine, so its rejection is
xlq\textquotesingle s engine rejecting a competitor\textquotesingle s
output, not a neutral arbiter; the one neutral engine measured,
LibreOffice, \emph{opens} the same openpyxl file. What is measured and
unconditional is that openpyxl drops \texttt{sharedStrings.xml} on every
fixture, and drops all VBA and all chart-support parts on the fixtures
that have them. Against every whole-file substrate, then, the surgical
writer is the only tool that leaves the parts an edit did not touch
exactly as it found them --- which is the property that makes the
boundary of §5 meaningful.

\subsection{4. Does It Work For an
Agent?}\label{4-does-it-work-for-an-agent}

A static fidelity benchmark (§3) shows the writer preserves parts when
\emph{we} drive it. It cannot show that the property survives contact
with a real agent --- an agent that chooses its own tool invocations,
reads its own results, and decides for itself whether it succeeded. This
section reports an interventional \emph{case study} --- n = 2 tasks, one
model, one trajectory per arm --- in which a real LLM agent completes a
genuine edit task, and the only manipulated variable is the tool it is
given to write the file (docs/AGENT-AB.md, benchmarks/agent\_ab.json).
It is demonstration-scale by design: it shows that the harm and its
prevention exist and are reproducible under a real agent, not that they
occur at any particular rate. This is the experiment the thesis always
demanded, and the one whose absence the earlier version of this work was
rejected for.

\subsubsection{Design}\label{design}

The agent is a real LLM subagent (Claude) with a bounded tool surface: a
shell and a Python interpreter, or the \texttt{xlq} binary. It is given
a plain-English edit task and told to complete and save it. It is
\emph{not} told which outcome is being measured, nor that fidelity is
scored. Two arms hold the agent, prompt, input file, and environment
fixed and vary only the write tool:

\begin{itemize}
\tightlist
\item
  \textbf{\texttt{openpyxl}} --- the status-quo path an unconstrained
  coding agent reaches for today, the one implicated in \#22044. The
  agent edits with the openpyxl library and saves.
\item
  \textbf{\texttt{xlq}} --- the format-aware boundary. The agent may
  mutate the workbook only through \texttt{xlq\ apply} (inspect →
  \texttt{patch.json} → apply).
\end{itemize}

Two real tasks, each a single \texttt{set\_cell} edit on a workbook
carrying a feature that whole-file rewriters destroy: \textbf{t1-vba}
(\texttt{macro.xlsm}, VBA project, change \texttt{Data!B3} 120 → 250)
and \textbf{t2-chart} (\texttt{pivot-chart.xlsx}, 2 charts + a pivot,
change \texttt{Sheet1!A2} 222 → 900). Each trial scores two things by
\emph{external check}, not by the agent\textquotesingle s own claim:
\textbf{task success} (did the target cell reach the requested value
\emph{and does the saved file still open?}) and \textbf{fidelity of
untouched content} (of the parts the edit did not touch, how many
survive byte-identical; which feature parts survive; which are dropped).
Because everything but the edit tool is held fixed, any difference in
outcome is attributable to the tool boundary.

\subsubsection{Per-task results}\label{per-task-results}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}llllllllll@{}}
\toprule\noalign{}
Task & Feature & Arm & Target & Task success & Reloads & Feature
byte-identical & Parts byte-identical & Parts dropped & Receipt \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
t1-vba & VBA & \textbf{openpyxl} & 250 ✓ & ✅ yes & ✅ yes & ✅ yes¹ &
\textbf{2 / 11} & 1 (\texttt{sharedStrings.xml}) & --- \\
t1-vba & VBA & \textbf{xlq} & 250 ✓ & ✅ yes & ✅ yes & ✅ yes &
\textbf{10 / 11} & 0 & ✅ rev-1 \\
t2-chart & 2 charts + pivot & \textbf{openpyxl} & \emph{load-error²} &
❌ \textbf{no} & ❌ \textbf{no} & ❌ no / ❌ no & \textbf{1 / 50} &
\textbf{13} & --- \\
t2-chart & 2 charts + pivot & \textbf{xlq} & 900 ✓ & ✅ yes & ✅ yes &
✅ yes / ✅ yes & \textbf{48 / 50} & 1 (\texttt{calcChain.xml}³) & ✅
rev-1 \\
\end{longtable}
}

¹ In this run the openpyxl agent volunteered \texttt{keep\_vba=True}, so
\texttt{vbaProject.bin} survived byte-identical --- the
\emph{charitable} openpyxl (threat T4). Even so it rewrote 9 of 11 parts
and silently dropped \texttt{sharedStrings.xml}. ²
\texttt{target\_cell\_value} came back as
\texttt{\textless{}load-error:\textquotesingle{}Chartsheet\textquotesingle{}\ object\ has\ no\ attribute\ \textquotesingle{}defined\_names\textquotesingle{}\textgreater{}}:
the saved file cannot be reopened, so the value is unverifiable because
the workbook is corrupt. The agent nonetheless \emph{reported success}
--- it read the raw XML, saw
\texttt{\textless{}c\ r="A2"\ t="n"\textgreater{}\textless{}v\textgreater{}900\textless{}/v\textgreater{}\textless{}/c\textgreater{}},
and concluded the edit landed. ³ \texttt{calcChain.xml} is a
recomputable formula-ordering cache, not a feature; its removal is
expected and lossless.

\textbf{The false-success finding.} The t2-chart / openpyxl trajectory
is the headline. The agent\textquotesingle s cell write \emph{did} land
--- the number 900 is in the XML --- but the workbook did not survive
the save: the two charts and the pivot were stripped of their
relationship parts, 13 of 50 parts were dropped, and only 1 of 50
remained byte-identical. When the agent\textquotesingle s own post-save
reload raised
\texttt{\textquotesingle{}Chartsheet\textquotesingle{}\ object\ has\ no\ attribute\ \textquotesingle{}defined\_names\textquotesingle{}},
it worked around the failed check by reading the raw sheet XML, saw its
target cell, and \textbf{reported success on a corrupt, non-reloadable
file}. This is \#22044 reproduced live in a single observed trajectory,
including the property that made \#22044 dangerous in the first place:
the corruption is invisible to the agent\textquotesingle s own success
criterion. We report it as one observed instance of the false-success
failure, not a frequency. In the paired trial the xlq-confined agent
completed the \emph{same task on the same file} with charts and pivot
byte-identical, 48 of 50 parts untouched, the file reloadable, and a
signed receipt (\texttt{rev\ 1}, base→result hash chain). \textbf{In
this trial the xlq-confined agent could not commit the \#22044
corruption even while attempting the identical edit} --- and the reason
is mechanical and holds for the tool generally, not just this
trajectory: \texttt{xlq\ apply} has no operation that rewrites the whole
container.

\subsubsection{Threats to validity}\label{threats-to-validity}

We state these plainly so the reader does not over-generalize
(docs/AGENT-AB.md §4):

\begin{itemize}
\tightlist
\item
  \textbf{T1 --- n = 2 tasks.} Two workbooks, two features. This
  demonstrates the harm and the prevention \emph{exist and are
  reproducible under an agent}; it is not a rate estimate. It does not
  claim "openpyxl corrupts X\% of workbooks."
\item
  \textbf{T2 --- single operation type.} Both tasks are one
  \texttt{set\_cell} each. Structural edits (insert/delete rows, add
  sheets, formula rewrites) are not exercised under the case study; the
  causal claim is scoped to value edits.
\item
  \textbf{T3 --- the "agent" is an LLM subagent with a bounded tool
  surface.} It is a real model making real tool calls, but not a full
  autonomous IDE agent with retrieval, retries, or user follow-up, and
  we measured a single trajectory per cell, not a distribution over
  agent skill. A more capable agent might notice and repair the openpyxl
  corruption; a less careful one would ship it.
\item
  \textbf{T4 --- the openpyxl arm is the \emph{charitable} openpyxl.} On
  t1 the agent chose \texttt{keep\_vba=True}. The realistic default is
  \texttt{keep\_vba=False}, which drops the VBA project entirely and
  silently (§3, E1) --- a strictly worse and more common outcome. Our
  openpyxl numbers therefore \emph{understate} the status-quo harm, not
  overstate it.
\item
  \textbf{T5 --- "reloads in engine" is checked against ironcalc}, the
  engine xlq vendors. The t2 openpyxl failure also reproduces under
  openpyxl\textquotesingle s \emph{own} reader (the
  \texttt{Chartsheet.defined\_names} load-error), so the failure is not
  an engine-choice artifact --- though a reader that tolerates missing
  chart \texttt{\_rels} might open the file in a degraded state rather
  than erroring.
\end{itemize}

\textbf{The causal claim, precisely scoped.} In these two paired trials,
holding the agent, task, input workbook, and environment fixed, routing
the edit through xlq instead of openpyxl changed the outcome from
"feature-destroying, non-reloadable, unattested" to "byte-identical on
all feature parts, reloadable, and receipted." On the chart/pivot task
this is the difference between reproducing \#22044 and being
mechanically unable to. The reason is mechanical, not statistical:
\texttt{xlq\ apply} rewrites only the parts an operation touches and
refuses whole-container rewrites, so the case study is a demonstration
that this mechanism survives contact with a real agent --- which the
static benchmark of §3 could not show --- while the two-trajectory
sample size is exactly why we call it a case study and not a rate.

\subsection{5. Threat Model}\label{5-threat-model}

A built write path invites a sharper objection than a read-only one:
even a surgical, receipted write is \emph{opt-in} --- nothing forces an
agent through xlq, so what exactly does it enforce? This section answers
precisely, stating what xlq guarantees, against whom, and where the
boundary is and is not (docs/THREAT-MODEL.md) --- and, in this version,
\emph{tests} the non-bypassability claim in a single adversarial-agent
trial rather than only arguing it from the harness design. One trial is
not a proof; it is one adversary who tried, failed to forge, and found a
real hole we then closed. The honest answer matches how the 2025--26
enforcement literature scopes its own boundaries, and we make the
parallel explicit rather than overclaim.

\textbf{The deployment: a confined agent.} xlq is an enforcement
boundary \emph{for an agent confined to it}. The concrete deployment is
the standard tool-sandbox model of the enforcement wave {[}IsolateGPT
2024; Progent 2025{]}: the agent harness grants the agent the
\texttt{xlq} tool and \textbf{no capability to write \texttt{.xlsx}
bytes by any other means} --- no openpyxl, no raw file write, no shell
that can. The shipped \texttt{skills/xlq/SKILL.md} plus a restricted
tool surface is that harness. Within it, \texttt{xlq\ apply} is the
agent\textquotesingle s \emph{only} path to modify a workbook, and every
modification is surgical (only sheet parts with an edited cell are
rewritten; charts, pivots, VBA, styles, external links byte-identical
--- §3), preview-gated (\texttt{-\/-dry-run} predicts affected cells and
new errors first), precondition-checked (\texttt{base\_hash} must match
under an advisory lock, no TOCTOU), coverage-gated (a write whose
predicted cached values are unreliable, or whose formulas use a function
the differential oracle found divergent, is refused --- §6, §7),
proof-carried and fidelity-enforced (before commit the output is
re-loaded and every predicted cell re-verified, \emph{and} every
non-edited part is byte-diffed against the input, aborting with
\texttt{fidelity\_violation} on any change --- §3), and receipted (a
hash-chained journal entry records base/result hash, ops, and actor;
out-of-band edits are detected, not silently absorbed). So an agent
confined to xlq \textbf{cannot reach the \#22044 corruption}: it has no
operation that rewrites the whole container or drops a chart/pivot/VBA
part. That is the enforcement claim, and §4 is its interventional
case-study evidence.

\textbf{What the boundary is \emph{not}.} It is not a guarantee against
an agent with unrestricted shell or raw filesystem access. Such an agent
can call openpyxl directly and corrupt the file --- exactly the status
quo --- and xlq does not, and cannot, prevent that; no library-level
tool can. Non-bypassability is a property of the \emph{harness} (which
tools it exposes), which xlq \emph{enables} but does not itself enforce.
We claim non-bypassability \emph{within the harness}, and say so, rather
than claim a guarantee against an adversary handed raw write access.
This mirrors the enforcement wave exactly: Progent\textquotesingle s
policy engine, IsolateGPT\textquotesingle s process isolation, and
MiniScope\textquotesingle s scoped tokens all bound an agent
\emph{routed through them}; none prevents an operator from also handing
the agent an unmediated capability. The boundary is the mediation
surface, and its value is that a harness \emph{can} confine an agent to
it --- for spreadsheets, previously it could not, because no safe write
primitive existed.

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
Actor & Can it corrupt via xlq? & Notes \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Careless agent confined to xlq & No & No whole-file-rewrite operation
exists; surgical by construction. \\
Buggy agent (wrong cell/formula) & Bounded & dry-run previews the
effect; \texttt{base\_hash} + receipts make every change attributable
and the rev-file makes it reversible. \\
Agent with raw shell / openpyxl & Yes (bypasses xlq) & Out of scope by
construction; the harness, not xlq, must withhold raw write. Stated as a
limitation. \\
Agent truncating/overwriting the file directly & Blocked within harness
& Not corruption \emph{via} xlq but blunt destruction; closed by a
read-only authoritative store + a receipt-gated broker (shown below in
an adversarial trial, NON-BYPASS.md). \\
Concurrent writers & Serialized & Advisory lock; the second writer gets
\texttt{lock\_held}. \\
Out-of-band edit between ops & Detected & Hash-chain
\texttt{external\_edit} + adoption marker. \\
\end{longtable}
}

\emph{Table 2. Threat actors and coverage (THREAT-MODEL.md).}

\textbf{Non-bypassability, tested adversarially --- not merely argued.}
The re-review\textquotesingle s standing objection was that this
enforcement claim was \emph{argued} from the harness design but never
\emph{tested} against an adversary who tries and fails to bypass it. We
ran a single adversarial trial (n = 1), and it both confirmed the
forgery claim and found a real hole we then closed (docs/NON-BYPASS.md);
we present it as one adversarial probe that surfaced and closed a
concrete gap, not as a bypass-resistance proof. A confined harness
(\texttt{env\ -i\ PATH=./confined-bin}) exposes exactly one executable
--- \texttt{xlq} --- plus \texttt{/bin/sh} builtins; \texttt{python3},
\texttt{cp}, \texttt{zip}, \texttt{unzip}, \texttt{dd}, \texttt{base64},
and every archiver or binary editor return "not found," verified. An
adversarial LLM agent, instructed to be impatient and to actively bypass
the safe path (rewrite the file with Python, unpack the zip, anything
faster), reported its result honestly:

\begin{itemize}
\tightlist
\item
  \textbf{Forgery --- blocked completely.} "I genuinely could not
  produce a modified-but-valid workbook by any route except xlq." With
  no interpreter and no archiver, there was no route to a
  \emph{valid-but-wrong} workbook; the \#22044 class --- a corrupt file
  that looks fine and is reported as success --- was unreachable.
\item
  \textbf{Destruction --- not blocked (the hole).} The working directory
  was writable and \texttt{:\ \textgreater{}\ target.xlsx} is a shell
  builtin, so the agent could \emph{truncate} the file --- no forgery,
  but blunt destruction, with no xlq and no receipt. The confinement
  prevented forgery, not destruction.
\end{itemize}

We report the hole as the rigor, not a weakness: an adversarial
experiment exists precisely to surface it, and the fix follows
first-principles from the finding (docs/NON-BYPASS.md §2). The
authoritative store is made read-only to the agent ---
\texttt{authoritative/book.xlsx} is \texttt{chmod\ 444} inside a
\texttt{chmod\ 555} directory, and the confined PATH has no
\texttt{chmod}, so the shell cannot restore write; truncation,
\texttt{echo\ \textgreater{}}, and new-file creation against it all
return permission denied, verified. The agent instead works on a
\emph{copy} in a writable \texttt{work/} dir via \texttt{xlq\ apply} (it
may freely destroy its own scratch copy --- harmless), and a
\textbf{receipt-gated broker} running \emph{outside} the confinement
promotes \texttt{work/book.xlsx} to \texttt{authoritative/} only if it
hashes to the last receipt\textquotesingle s \texttt{result\_hash}
\emph{and} that receipt chains from the authoritative
file\textquotesingle s current hash, rejecting anything not
receipt-backed or not a clean successor. End-to-end, measured: the
confined agent applied its edit (\texttt{rev\ 1},
\texttt{verified.all\_landed\ =\ true}, fidelity 48/50), the broker
verified the chain (\texttt{base\ 35010189…\ →\ result\ abbed433…}) and
promoted, and the authoritative workbook afterward carried \textbf{A2 =
900} with \textbf{10 of 10 chart parts byte-identical} to the original
and the receipt chain intact --- changed only through a receipted,
proof-carried, fidelity-preserving xlq operation, with no route for the
agent to forge or destroy it.

The honest scope sentence, which we state in the paper rather than bury:
\emph{xlq is an enforcement boundary for an agent whose harness confines
its workbook writes to the xlq tool; within that confinement it makes
the \#22044 corruption unreachable and every change previewed,
attributable, and reversible --- now tested in one adversarial trial,
forgery blocked and the one destruction hole found and closed. It does
not defend a workbook against an agent granted an unconfined shell with
an interpreter and a writable authoritative file --- that is the
harness\textquotesingle s responsibility, and the two-tier
read-only-store-plus-broker mechanism above is exactly how a harness
discharges it, stated as a scoping assumption rather than a guarantee.}

\subsection{6. Reading Without Writing: The Census and Coverage
Honesty}\label{6-reading-without-writing-the-census-and-coverage-honesty}

The write path of §3 is only trustworthy if the read that precedes it
is: an agent must see the workbook to build a patch, the dry-run must
predict an edit\textquotesingle s effect, and the boundary must know
when it \emph{cannot} predict. This section describes the read layer ---
a privacy-safe structural census and a per-file coverage report ---
carried over from the read-only system and now load-bearing for the
write. Both are reads that are side-effect-free by construction; the
mutating surface is \texttt{apply} alone (§3).

\subsubsection{A privacy-safe structural
census}\label{a-privacy-safe-structural-census}

\texttt{xlq\ inspect} produces a \emph{census}: a structural summary
that answers "what does this workbook need from an engine?" without
answering "what is in it?" It reports sheet dimensions and cell/formula
tallies, per-literal counts of stored errors, the call-site frequency of
each Excel function, which OOXML parts are present, and the
file\textquotesingle s SHA-256 --- and, by specification, never a cell
value, never a fragment of a formula body, and never a full filesystem
path. The invariant is regression-tested against a sentinel workbook
whose cells carry \texttt{SECRET\_VALUE\_XYZ} and
\texttt{SECRET\_FORMULA\_LIT}: the census is serialized in normal and
redacted mode and asserted to contain neither marker nor the
file\textquotesingle s parent directory. Taking the format seriously
forces one distinction a format-blind summarizer would miss: a called
name that is neither a canonical Excel function nor known to the engine
--- a VBA/XLL user-defined function, an add-in call, a \texttt{LAMBDA}
through a defined name --- is treated as \emph{user data}, because such
names routinely encode deal terms and client identities (a call site in
our tests is \texttt{DealMargin\_AcmeCorp}); these are reported under a
separate \texttt{user\_defined\_calls} member, are redactable, and never
leak into the function tallies.

The census is also what makes the artifact legible to an agent at a
fraction of the token cost of dumping cells. On a workbook of roughly
100k formulas (\texttt{perf-large.xlsx}), the census was \textbf{999
bytes} against a \textbf{7,239,303-byte} naive full-cell JSON dump --- a
\textbf{7,246.5× reduction} (results.json); on the small
\texttt{branch-consolidation.xlsx}, 1,800 B versus 51,107 B (28.4×).
This is not comparable to prompt-side content-compression schemes such
as SpreadsheetLLM {[}Dong et al. 2024{]}, which compress cell
\emph{contents} for understanding and question-answering: the census
discards contents entirely and keeps only the structure a
mutation-safety decision needs.

\subsubsection{Coverage honesty: three numbers, not
one}\label{coverage-honesty-three-numbers-not-one}

A formula runtime is conventionally described by one number: how many of
Excel\textquotesingle s functions it "supports." That fuses two
questions the owner of a workbook needs answered apart --- does the
runtime \emph{recognize} this function, and can it \emph{compute the
function\textquotesingle s value from the workbook\textquotesingle s own
data}? --- and has no vocabulary for the case where the correct answer
is a specific documented error. We replace the bit with a partition.
Against a pinned catalog --- the 522 worksheet functions in
Microsoft\textquotesingle s alphabetical function list, fetched
2026-07-02 (\texttt{benchmarks/excel-functions.txt}) --- the vendored
engine measured (\texttt{benchmarks/coverage.json}):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}llr@{}}
\toprule\noalign{}
\# & Measure & Count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & \textbf{Recognized} --- the name parses and dispatches to defined
behavior, never a parser unknown-name rejection & \textbf{522 / 522} \\
2 & \textbf{Locally evaluable} --- full semantics computed from the
workbook\textquotesingle s own data & \textbf{505} \\
3 & \textbf{Policy-limited} --- recognized and argument-validated, then
the documented desktop-Excel refusal literal, because the true value
depends on something the runtime does not have & \textbf{17} \\
\end{longtable}
}

\emph{Table 3. The three-number coverage taxonomy (coverage.json). 505 +
17 = 522.}

The numbers are exhaustive and disjoint over the recognized set. A guard
states what number 2 does \emph{not} mean: \emph{locally evaluable is
not correct.} Evaluability is coverage --- the engine will compute a
defined value from workbook data --- whereas correctness, whether that
value matches Excel, is the oracle\textquotesingle s burden and is
measured separately in §7. The 505 must never be read as "505 correct."
The 17 policy-limited functions decompose, on the same honesty the
taxonomy exists to enforce, into three reasons a reader would act on
differently (coverage.json):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lrll@{}}
\toprule\noalign{}
Reason class & n & Example functions & Literal \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Capability-limited} & 14 & WEBSERVICE \texttt{\#VALUE!}; RTD
\texttt{\#N/A}; STOCKHISTORY / DETECTLANGUAGE / TRANSLATE / COPILOT /
IMAGE \texttt{\#CONNECT!}; 7 CUBE functions \texttt{\#NAME?} & offline
desktop-Excel literal after local argument validation \\
\textbf{Policy-blocked} & 2 & CALL, REGISTER.ID (XLM/DLL native-code
entry points) & \texttt{\#BLOCKED!} --- refused for security (memo
§16) \\
\textbf{Context-precondition} & 1 & GETPIVOTDATA (needs a rendered
PivotTable the engine does not model) & \texttt{\#REF!} \\
\end{longtable}
}

\emph{Table 4. The 17 policy-limited functions, decomposed by reason
(coverage.json).}

Coverage honesty is a \emph{per-file} property, and this is what ties it
to the write path. When a file uses a policy-limited name,
\texttt{xlq\ inspect} and \texttt{xlq\ calc} report it in the
census\textquotesingle s \texttt{policy\_limited\_functions} bucket
(name → literal) and set \texttt{coverage.reliable:\ false} --- the
stored values genuinely cannot be \emph{verified} locally --- while
\texttt{unsupported\_functions} stays empty, because "the engine does
not know this name" would be a lie. The dry-run (§3) reuses exactly this
machinery: if a volatile or policy-limited function lies in an
edit\textquotesingle s affected dependency graph, the
prediction\textquotesingle s coverage flag degrades and the write is
refused, so the boundary never commits an edit whose predicted
downstream values it cannot stand behind. A companion gate driven by the
differential oracle\textquotesingle s confusion matrix (§7) refuses the
write on the same principle when a formula uses a function the engine is
known to compute wrong --- the two gates together bound both what the
engine \emph{cannot} evaluate and what it evaluates \emph{incorrectly}.
The three numbers were reached by measuring before building: the pinned
IronCalc 0.7.1 release recognized 345/522 (66.1\%), but upstream
\emph{master} had already added \textasciitilde148 functions, so reading
the diff --- not writing code --- moved recognition to 494/522 (94.6\%);
three residual local patches (AGGREGATE, ENCODEURL, HYPERLINK) reached
497, and a final milestone implemented eight more with local semantics
(FILTERXML, EUROCONVERT, DBCS, JIS, BAHTTEXT, PHONETIC, GROUPBY,
PIVOTBY), taking locally evaluable to 505 and closing recognized to
522/522 (\texttt{docs/COVERAGE.md}). This makes claim \textbf{C2}: a
formula runtime can recognize the full Excel catalog with defined,
documented behavior and report that coverage \emph{honestly}, as a
three-number partition with a per-function literal for every
non-evaluable case, rather than as a single supported-count that
conflates recognition, evaluability, and correctness.

\subsection{7. Documentation-Arbitered Differential
Validation}\label{7-documentation-arbitered-differential-validation}

The census (§6), the coverage report (§6), and above all the dry-run
prediction that gates every write (§3) are only as trustworthy as the
formula engine beneath them: a vendored engine that computes a wrong
value will have the runtime faithfully report a wrong value, and predict
a wrong downstream effect. Correctness is a separate burden ---
contribution C3 --- and it cannot be discharged by assumption. Nor can
it be discharged the way a compiler\textquotesingle s or a SQL
engine\textquotesingle s can, because there is no mechanized
specification of spreadsheet formula semantics: ECMA-376 standardizes
the OOXML container, not the numerical behavior of the functions inside
it, and OpenFormula governs ODF, not Excel\textquotesingle s dialect. So
the engine was validated the way compilers were before verified
reference semantics existed {[}Yang et al. 2011{]}: by differential
testing. Each of 1,659 hand-authored cases across 492 functions was run
through two independent engines --- the vendored IronCalc (master
\texttt{e50ccea8}) and LibreOffice 24.8.7.2 headless --- and every
divergence was treated as a signal to investigate, not a verdict. Of the
1,659 cases the engines agreed on 1,273, disagreed on a value in 85,
both raised an error in 196, and hit 105 cases LibreOffice does not
implement (15 of which IronCalc also errored on); no case crashed either
engine (\texttt{engine\_error\ =\ 0}, benchmarks/agreement.json).

\subsubsection{The disagreement matrix}\label{the-disagreement-matrix}

The scientific object is not the agreement count but the 85 cases in
which both engines returned a value and the values differed. Each was
adjudicated by hand against Excel\textquotesingle s documented behavior,
using a three-way verdict discipline {[}Liu et al. 2024{]} --- a
divergence is a \emph{bug}, an \emph{ambiguity}, or \emph{undecidable},
never auto-blamed on one side (benchmarks/triage-analysis.md):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}ll@{}}
\toprule\noalign{}
Excel-arbitrated verdict & Count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
IronCalc wrong & 24 \\
LibreOffice wrong & 41 \\
Both wrong & 7 \\
Spec-ambiguous & 13 \\
Undecidable here & 0 \\
\end{longtable}
}

\emph{Table 5. Excel-arbitrated triage of the 85 value-vs-value
disagreements (triage-analysis.md).}

Seventy-two of the 85 (85\%) received a decidable,
documentation-grounded fault assignment. Two properties matter more than
any cell. First, the harness indicts the engine the runtime actually
ships \textbf{24 times} --- a rigged harness does not fault its own
engine two dozen times. Second, on the 18 core-semantics cases with zero
ambiguity the split is nearly even (8 IronCalc-wrong to 10
LibreOffice-wrong), while on the 21 financial cases it is lopsided the
other way (13 LibreOffice-wrong to 3 IronCalc-wrong, dominated by
day-count and duration errors): the two engines fail in \emph{different
directions}, which is precisely why comparing them was informative.
Concrete filable defects resolve from the verdicts --- IronCalc:
\texttt{CONVERT(68,"F","C")} returned 19.65 where Excel returns 20,
\texttt{ROW} over a range returned a scalar instead of the spilled
array, \texttt{SUMPRODUCT} did not coerce boolean arrays, the
\texttt{A}-family aggregates ignored text and boolean cells,
\texttt{TRIM} did not collapse internal whitespace, \texttt{PRICE} at
basis 3 returned exact par; LibreOffice: \texttt{POWER(0,0)} returned 1
where Excel returns \texttt{\#NUM!} (and IronCalc matched Excel), a
four-function cluster treated boolean cells as plain numbers, four
Treasury-bill functions used a wrong day count, and
\texttt{DURATION}/\texttt{MDURATION} diverged from documented Macaulay
duration (which IronCalc reproduced to 1e-9).

\textbf{Concordance is not accuracy.} Only beneath the matrix does the
aggregate belong. On the 1,311 numerically-comparable cases the engines
matched on 1,273 --- \textbf{97.1\%} (935 exact, 338 within tolerance);
across all 1,358 both-value cases, 93.7\% (agreement.json). Two engines
reading the same ECMA-376 and vendor prose can converge on the same
wrong behavior, so agreement bounds the \emph{disagreement} count from
below, not the \emph{error} count from above. LibreOffice is a reference
peer, not ground truth; this paper never claims "IronCalc is 97.1\%
correct." The only correctness statements it makes are the 72 arbitrated
per-case verdicts, each traceable to Excel documentation.

\subsubsection{The confusion matrix as the write-reliability
gate}\label{the-confusion-matrix-as-the-write-reliability-gate}

The oracle is not only after-the-fact validation; its confusion matrix
is \emph{load-bearing for write safety}, and wiring it into the write
path answers the sharpest systems-security finding of the re-review.
That finding: the surgical writer persists engine-computed cached values
on the sheets it rewrites, and the oracle itself found the engine
computes some functions wrong --- which reopens the value-corruption
class the boundary exists to close and appears to contradict the
engine-independence of the fidelity property. The fix folds the oracle
directly into the write path. Every IronCalc-wrong verdict above names a
function; a real \texttt{xlq\ apply} refuses to commit
(\texttt{coverage\_unreliable},
\texttt{oracle\_divergent\_functions:\ {[}...{]}}) when any cell in the
edit\textquotesingle s affected dependency graph uses a function on that
divergence list --- CONVERT, TRIM, ROW, SUMPRODUCT, MAXA/MINA/STDEVA,
SECOND, PRICE, PERCENTRANK.EXC. So xlq never persists a computed cache
it has documented reason to believe is wrong. The gate is scoped
precisely: it fires only on \emph{engine-computed} formula caches,
because those are the values the oracle can indict --- a literal
\texttt{set\_cell} writes exactly what the user typed and is
engine-independent, so it is never gated. This also retires the
re-review\textquotesingle s "loosely coupled" critique: the differential
oracle and the write boundary are no longer two adjacent components but
one mechanism, the confusion matrix deciding which computed values the
writer is permitted to persist.

\subsubsection{E3: an independent financial cross-check retires
circularity for two verdicts and downgrades a
third}\label{e3-an-independent-financial-cross-check-retires-circularity-for-two-verdicts-and-downgrades-a-third}

The largest fault asymmetry --- 13 LibreOffice-wrong to 3 IronCalc-wrong
on the financial functions --- rested, in the earlier work, entirely on
author-written Python reimplementations of the day-count and pricing
formulas: a single in-house third implementation, corroborating but not
independent. To retire that circularity, we cross-checked the financial
verdicts against an \emph{independent} reference ---
\texttt{numpy-financial} plus documented Treasury-bill formulas,
implemented separately from the triage reimplementation
(benchmarks/financial\_crosscheck.json). The result both strengthens and
corrects the oracle, and we report the correction as prominently as the
confirmations:

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}llllll@{}}
\toprule\noalign{}
Function (case) & Independent reference & IronCalc & LibreOffice &
Original triage & Cross-check outcome \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
TBILLPRICE (184d, 0.045) & 97.7 & 97.7 & 97.7375 & LibreOffice wrong
(DSM=181) & \textbf{CONFIRMS IronCalc}; LO uses DSM=181 \\
TBILLYIELD (184d, 98.5) & 0.0297947 & 0.0297947 & 0.0302886 &
LibreOffice wrong (DSM=181) & \textbf{CONFIRMS IronCalc}; LO uses
DSM=181 \\
TBILLEQ (184d, 0.045) & 0.0466813 & 0.0466902 & 0.0466812 & LibreOffice
wrong & \textbf{CONTRADICTS}: independent value matches LibreOffice, not
IronCalc \\
RATE(10,−100,1000) & −9.49e-11 & −3.35e-11 & 6.12e-11 & Spec-ambiguous &
CONFIRMS both ≈ 0 (true rate 0) \\
\end{longtable}
}

\emph{Table 6. Independent financial cross-check
(financial\_crosscheck.json).}

For \textbf{TBILLPRICE} and \textbf{TBILLYIELD}, the independent
reference reproduces IronCalc\textquotesingle s value to the reported
precision and confirms that LibreOffice used a 181-day count where the
documented convention gives 184. Those two verdicts no longer rest on
our own reimplementation; the circularity objection is retired for them.
For \textbf{TBILLEQ}, the cross-check \emph{contradicts the original
triage}: the independent bond-equivalent form matches
LibreOffice\textquotesingle s \texttt{0.0466812}, not
IronCalc\textquotesingle s \texttt{0.0466902}. We therefore
\textbf{downgrade that verdict from "LibreOffice wrong" to
UNDECIDABLE-HERE pending a licensed Excel} --- the independent
bond-equivalent formula may itself differ from Excel\textquotesingle s
exact algorithm, so neither engine can be faulted on the present
evidence. We state this as a \emph{strength of the method, not a defect
hidden}: an oracle whose own independent check can overturn one of its
verdicts is an oracle that is falsifiable, and the TBILLEQ downgrade is
exactly the kind of self-correction a hand-adjudicated triage most
needs. It also sharpens the honest scope of the financial row: two of
its verdicts are now independently corroborated, one is withdrawn, and
the cleanest remaining strengthening --- the one this cross-check cannot
supply --- is Excel-the-binary as a third differential peer, unavailable
on the Linux host and stated as future work.

\subsubsection{What is, and is not, new
here}\label{what-is-and-is-not-new-here}

None of the \emph{method} is new: generating inputs in a well-defined
subset and cross-voting independent implementations is the
differential-testing playbook {[}Yang et al. 2011{]}, and arbitrating
divergences against a prose standard is what CSmith does against the C
standard. The contribution claimed for C3 is (a) the \emph{domain} ---
spreadsheet formula engines, which unlike compilers {[}Yang et al.
2011{]}, SQL engines {[}Rigger and Su 2020{]}, and JS engines {[}Park et
al. 2021; SmartOracle 2026{]} have no mechanized specification to build
a stronger oracle from; (b) an explicit, artifact-backed arbitration
protocol for that spec-less setting, now including an independent
cross-check that overturned one of its own verdicts; and (c) the
concrete bugs it surfaced in both engines. Two blind spots are named
rather than smoothed: the 1,659 cases are hand-authored at roughly 3.4
per function, so 85 disagreements is a \emph{lower bound} on divergence;
and the strongest arbiter, Excel-the-binary, entered only as
documentation because no licensed Excel was available. This is what
earns the runtime\textquotesingle s honesty: the
engine\textquotesingle s outputs --- and the dry-run predictions built
on them --- are trusted because they are differentially validated with
documentation arbitration and, for the financial functions, an
independent cross-check, not because IronCalc is assumed correct.

\subsection{8. AXLE-bench: Measuring the Omitted
Dimension}\label{8-axle-bench-measuring-the-omitted-dimension}

Prior agent benchmarks score whether the task cells came out right; none
reports whether the file survived. AXLE-bench (Agent-safe eXcel Layer
Evaluation) is a consolidated suite that makes the fidelity gap of §2 a
first-class measurement axis, beside the four dimensions on which a
spreadsheet tool is ordinarily judged. Following
OS-Harm\textquotesingle s taxonomy-before-benchmark discipline
{[}OS-Harm 2025{]}, the axes are fixed first and every tool instantiated
against them symmetrically. Every number traces to a JSON artifact in
\texttt{benchmarks/}, and one meta-runner reproduces them on a single
commodity machine (Intel i7-8700K, 12 logical cores, Linux). The five
axes are: \textbf{Correctness} --- the differential oracle and its
disagreement triage, headline being the confusion matrix of §7 (now
load-bearing for write safety: it gates which engine-computed caches a
real apply may persist), not the 97.1\% concordance; \textbf{Fidelity}
--- the omitted dimension, now measured two ways: the whole-file re-save
tiers of §2 (101,961 blanked by openpyxl, 90,448 drifted by LibreOffice)
\emph{and} the per-file surgical-preservation results of §3 (E1) on real
charts/pivots/VBA, plus the interventional case study of §4 (E2);
\textbf{Efficiency} --- median-of-3 warm wall time on the
\textasciitilde100k-formula fixture (\texttt{xlq\ calc}, which loads,
fully recalculates, audits stored-vs-recomputed, hashes, and reports, at
1.264 s; isolated engine load 0.599 s;
\texttt{libreoffice\ -\/-convert-to} 1.494 s; openpyxl
\texttt{load\_workbook} 0.875 s with recalc n/a --- it has no formula
engine); \textbf{Agent-ergonomics} --- the census token ratio (999 B vs
7,239,303 B, 7,246.5×); and \textbf{Catalog} --- the three-number
coverage (522/522 recognized, 505 evaluable, 17 policy-limited).

\textbf{Instrument validation as a co-headline.} The measuring
instruments are themselves fallible, so their limits are reported beside
their outputs. The correctness axis has no live Excel oracle; the 1,659
cases are hand-authored at \textasciitilde3.4 per function, making the
disagreement rate a \emph{lower bound}; and, new in this version, an
independent financial cross-check \emph{overturned one of the
oracle\textquotesingle s own verdicts} (TBILLEQ, §7) --- the sharpest
instrument-validation point available, because the tool caught its own
error. The fidelity axis compares zip central-directory part names,
sizes, and CRC32s, a byte-level basis that is exactly why cosmetic drift
(LibreOffice\textquotesingle s 90,448) has to be separated from
consumer-breaking loss (openpyxl\textquotesingle s 101,961) by hand, and
why the surgical writer\textquotesingle s "byte-identical elsewhere"
property is a meaningful signal rather than noise. An earlier
\texttt{xlq\ diff} once reported one change on an openpyxl re-save while
442 formula results had silently vanished --- the exact damage the
boundary exists to catch, hiding inside the measuring tool --- which is
why the \texttt{cached\_value} change-kind exists. We report these blind
spots because the property the suite is built to measure is a
tool\textquotesingle s willingness to report its own.

\textbf{The comparison, in one claim.} No other tool in the matrix wraps
agent writes in preview → apply → receipt → revision, and none reports
its own per-artifact blind spots --- not openpyxl, not LibreOffice, not
the higher-star MCP and CLI wrappers built on openpyxl underneath
(verified against each tool\textquotesingle s documentation,
\texttt{docs/BASELINE.md}). And now, crucially, no other tool leaves the
parts an edit did not touch byte-identical: the surgical writer is alone
in the "fidelity by construction" cell, which the whole-file substrates
cannot occupy however careful their save. That gap --- not raw speed ---
is the dimension AXLE-bench exists to measure and to keep honest.

\subsection{9. Related Work}\label{9-related-work}

We place related work after the mechanism so the reader meets the
boundary before the literature it draws on. We did not invent receipts,
transactional sandboxing, or differential testing. The claimed
contribution is narrower: making an agent \emph{write} boundary aware of
its target\textquotesingle s \emph{format fidelity} --- surgical, with a
fidelity property enforced and checked per apply, and demonstrated under
a real agent --- and validating a formula engine by
documentation-arbitered differential testing in a domain with no
mechanized specification.

\textbf{Runtime enforcement for LLM agents.} A 2024--26 wave established
that deterministic runtime enforcement outperforms prompt- and
heuristic-level defenses. Progent {[}Progent 2025{]} reports 0\% attack
success under deterministic policies against 39.9\% for prompt-based
mitigation; AgentSpec {[}AgentSpec 2025{]} compiles runtime rules from a
DSL; MiniScope {[}MiniScope 2025{]} derives least-privilege OAuth
scopes; IsolateGPT {[}IsolateGPT 2024{]} enforces process isolation; a
transactional sub-line makes writes reversible --- ACID-snapshot
sandboxing {[}Fault-Tolerant Sandboxing 2025{]} reports 100\% filesystem
rollback, and semantic/compensating-transaction systems {[}Cordon 2025;
SAFEFLOW 2025{]} extend it to multi-step actions. We adopt the shared
finding that a runtime is the only place a guarantee can live, and we
scope our boundary the way this wave scopes its own --- non-bypassable
\emph{within the harness}, not against an agent handed raw capability
(§5). The shared blind spot we differentiate against is that every one
of these systems is \emph{format-blind}: it can gate a filesystem write
or a shell command, but none understands the artifact it guards, so none
can answer \emph{will this edit corrupt the workbook\textquotesingle s
semantics?} Our write layer is not novel for being atomic; its
differentiators are that it is \emph{format-aware and built} --- the
surgical writer preserves parts by construction (§3), and the
dry-run\textquotesingle s prediction is \emph{semantic} (affected cells,
new formula errors, watched values), which requires the format-aware
engine the enforcement wave does not carry.

\textbf{Receipts and provenance.} Cryptographic action receipts and
tamper-evident logs are established: Notarized Agents {[}Notarized
Agents 2026{]} issues receiver-attested receipts for agent actions;
forward-secure MAC logging {[}Nitro 2025{]}, Certificate Transparency
{[}Certificate Transparency 2022{]}, and W3C PROV / PROV-AGENT {[}Moreau
et al. 2013{]} supply the provenance ancestry. Our receipt journal is a
domain instantiation --- a single-writer local sidecar with a hash chain
and out-of-band edit-adoption semantics, targeting custody of a file a
human co-edits --- now \emph{exercised} by the built \texttt{apply} path
(a receipt per revision in §3--§4) rather than specified only.

\textbf{Differential testing.} The oracle-construction menu is mature:
cross-vote generation in a well-defined subset {[}Yang et al. 2011{]},
metamorphic EMI {[}Le et al. 2014{]}, a tool-built reference evaluator
as arbiter {[}Rigger and Su 2020{]}, a mechanized specification as the
(N+1)-th implementation {[}Park et al. 2021{]}, and LLM-mediated triage
against a normative spec {[}SmartOracle 2026{]}, with Liu et al. {[}Liu
et al. 2024{]} contributing the three-way verdict vocabulary our triage
adopts. Each spec-as-arbiter technique presupposes a mechanized or
normative specification the spreadsheet domain lacks. Our instantiation
contributes documentation-arbitered triage over 1,659 curated cases, now
with an independent financial cross-check that corroborated two verdicts
and overturned a third (§7); while cross-engine spreadsheet-accuracy
comparisons against certified reference values exist {[}Almiron et al.
2010; McCullough 2008{]}, no prior work applies generative differential
testing with automated documentation-grounded triage to spreadsheet
formula engines.

\textbf{Spreadsheet research and LLM spreadsheet agents.} The motivation
bedrock is replicated error-pervasiveness work: Panko {[}Panko 2008{]}
found errors in 94\% of 88 audited sheets, and Hermans and Murphy-Hill
{[}Hermans and Murphy-Hill 2015{]} found Excel errors in 24\% of
formula-bearing Enron sheets, with 75\% using only the top-15 functions.
The EUSES corpus {[}Fisher and Rothermel 2005{]} and ExceLint {[}Barowy
et al. 2018{]} are classical ancestors of the structural census. On the
agent side, SheetCopilot {[}Li et al. 2023{]}, SheetAgent {[}SheetAgent
2025{]}, SpreadsheetBench {[}SpreadsheetBench 2024; SpreadsheetBench
2026{]}, and Finch {[}Finch 2025{]} all score task-answer cells by exact
match; none measures file corruption or preservation of untouched
content, SheetCopilot\textquotesingle s checker skips unflagged
properties by design, and only OS-Harm {[}OS-Harm 2025{]} measures agent
side effects at all --- and not file fidelity. The census occupies a
third representation design point distinct from model-side encoders
{[}Wang et al. 2021{]} and prompt-side lossy compression for
understanding {[}Dong et al. 2024{]}: a tool-side, structure-only
summary for mutation safety.

\textbf{Pista --- the closest prior.} Pista {[}Pista 2026{]}, with
Gulwani among its authors, is the nearest system: step-level human
oversight of a spreadsheet agent with model-generated explanations of
each action. Its guarantees are, by its authors\textquotesingle{}
account, "solely system prompt instructions," it neither verifies that
an explanation matches the executed snippet nor measures whether the
file survived, and it wants "programmatic controls" as future work. This
paper walks through that door with a machine-checked, format-aware
boundary that operates unattended, and it supplies what a synchronous
human auditor cannot certify: that the artifact the user is accountable
for is still byte-for-byte intact where the edit did not reach.

\subsection{10. Discussion}\label{10-discussion}

\subsubsection{The journey: from a read-only boundary that was rejected
to a built write path with an enforced fidelity
property}\label{the-journey-from-a-read-only-boundary-that-was-rejected-to-a-built-write-path-with-an-enforced-fidelity-property}

The honest history of this work is its strongest argument. An earlier
version diagnosed the fidelity gap correctly, built a read-only
inspector that could observe the \#22044 damage, and specified --- but
did not build --- the write path. A unanimous program-committee
rejection (mean −1.83) named three fatals, and each is now backed by a
built artifact rather than a specification. The "enforcement boundary"
was unbuilt, read-only, enforcement-by-absence of a write path; it is
now a built surgical write primitive with a fidelity property that holds
by construction \emph{and is enforced and checked per apply} --- each
write re-loading its own output to confirm the predicted edit landed and
byte-diffing every non-edited part, aborting with
\texttt{fidelity\_violation} on any change before committing (§3, 128
tests) --- so the word "enforcement" is earned rather than reached for.
There was no threat model and no account of opt-in; there is now an
explicit one whose non-bypassability was \emph{tested in a single
adversarial trial} in which an adversarial agent could not bypass the
confined harness --- forgery blocked, and the one destruction hole it
found closed by a read-only authoritative store and a receipt-gated
broker --- rather than merely argued (§5). No agent was ever run; in a
single observed trajectory a real LLM agent has now reproduced the
\#22044 corruption live through the status-quo tool --- including the
false-success failure where it reported success on a non-reloadable file
--- and been shown mechanically unable to commit it when confined to xlq
(§4). We keep these claims at the scale of their evidence: a case study
of two tasks, one adversarial trial, four fidelity fixtures. The
property that ties them together is that the write does \emph{not}
rewrite the whole file: because the surgical writer copies un-edited
parts byte-for-byte and the apply refuses to commit if any un-edited
part moved, an agent confined to it \emph{cannot} reach the failure
mode, and the reason is mechanical rather than a matter of the
agent\textquotesingle s care.

\subsubsection{The bug in our own tool}\label{the-bug-in-our-own-tool}

The finding that most validated the thesis was in our own hands. On a
re-saved copy of \texttt{branch-consolidation.xlsx}, \texttt{xlq\ diff}
once reported \emph{one change} while 442 formula results had silently
vanished --- an openpyxl round-trip had blanked every
\texttt{\textless{}v/\textgreater{}} cache, yet the diff, comparing
formatted display strings, saw only the single cell whose rendered text
moved and pronounced the file intact. The tool built to make silent
corruption visible was itself rendering corruption invisible. Surface
verification, not code review, caught it --- running the tool end to end
and reading its output against the file, the same discipline the
boundary imposes on an agent --- and the fix added the recompute-aware
\texttt{cached\_value} change-kind. The TBILLEQ downgrade in §7 is the
same discipline applied to the oracle: an independent cross-check
overturning one of our own verdicts is the strongest evidence that the
method is falsifiable rather than self-flattering.

\subsubsection{Load-bearing assumptions, stated
plainly}\label{load-bearing-assumptions-stated-plainly}

\emph{The fidelity property is by construction, enforced and checked per
apply, and corroborated by measurement.} Each write re-loads its own
output and confirms the predicted edit landed, and byte-diffs every
non-edited part and aborts on any change --- so both the edited half
("the written file verifiably computes it") and the
byte-identical-elsewhere half ("the un-edited parts are verifiably
untouched") are checked at write time, not only measured on the
fixtures. The single residual that still rests on the
writer\textquotesingle s control flow is \emph{minimality} --- that a
rewritten sheet had to be rewritten --- which the byte-diff cannot
establish, and a reviewer should read that one half as the assertion it
is (§3). \emph{IronCalc is still used for the dry-run prediction} ---
but it never touches the written file; the write is pure zip surgery, so
engine-fidelity limits affect the \emph{prediction quality} (flagged by
coverage-honesty and the coverage gate), not the preservation guarantee,
which is engine-independent --- and, for computed caches, the
engine\textquotesingle s reach is bounded further by the oracle
write-gate, which refuses to persist an engine-computed value whose
function the differential oracle found divergent (§7). \emph{Engine
trust is differential-earned, not assumed}, and for the financial
functions now independently cross-checked, with the one overturned
verdict reported. \emph{Non-bypassability is a property of the harness},
which xlq enables but does not itself enforce; we claim it within the
harness, tested it in one adversarial trial, and say exactly where it
stops.

\subsection{11. Limitations}\label{11-limitations}

Each limitation is tied to a specific mechanism; where it bounds a
headline number, we name the number and direction.

\emph{Small fidelity corpus; E1 is four hand-picked fixtures.} The
per-file preservation results (§3) are a demonstration on four workbooks
chosen to carry charts, pivots, and VBA, not a corpus-wide rate. The
byte-diff establishes only the byte-identical-elsewhere half; the
necessity of each rewritten sheet rests on the writer\textquotesingle s
construction. The LibreOffice column is a \texttt{-\/-convert-to}
re-save proxy, an \emph{upper bound on re-save churn}, not a
like-for-like single-cell edit.

\emph{The agent case study is two tasks, one model, one trajectory per
arm.} E2 (§4) is an interventional case study that demonstrates the harm
and its prevention exist and are reproducible under a real agent; it is
not a corruption rate and implies no generality. Both tasks are single
\texttt{set\_cell} edits --- structural mutations (row insert/delete,
formula rewrites, sheet add) are unexercised --- the "agent" is a
bounded LLM subagent rather than a full autonomous IDE agent, one
trajectory was observed per arm rather than a distribution over agent
skill, and the openpyxl arm is the \emph{charitable} openpyxl (the
realistic \texttt{keep\_vba=False} default is strictly worse), so the
status-quo numbers understate the harm.

\emph{The dry-run\textquotesingle s prediction depends on IronCalc.} The
write preserves parts regardless of the engine, but the \emph{predicted}
affected cells, new errors, and coverage flag are only as good as the
vendored engine; coverage-honesty flags where they cannot be trusted,
but a wrong-but-confident engine value would produce a
wrong-but-confident prediction that the coverage gate does not catch.

\emph{The oracle is hand-authored and lacks a live Excel; one financial
verdict is now undecidable.} The 1,659 cases at \textasciitilde3.4 per
function make the disagreement rate a lower bound, adjudication was by
hand without a second blind rater, and Excel entered only as
documentation. The independent cross-check retired circularity for
TBILLPRICE and TBILLYIELD but \textbf{downgraded TBILLEQ to undecidable
pending a licensed Excel} --- the single clearest remaining
strengthening of the oracle. Two residual weaknesses persist: the suite
is a hand-authored conformance set, not a generator (so coverage of each
function\textquotesingle s input space is shallow), and
E3\textquotesingle s cross-check still reimplements the same documented
formulas rather than consulting a fully external binary --- a licensed
Excel as a third differential peer is the strengthening neither the
cross-check nor the divergence gate can supply on the Linux host.

\emph{Non-bypassability was tested in one adversarial trial (n = 1), and
is still scoped to the harness.} In that single trial an adversarial
agent could not bypass the confined harness --- forgery blocked, and the
one destruction hole found and closed with a read-only authoritative
store and a receipt-gated broker (§5) --- but one trial is a probe, not
a bypass-resistance proof, and xlq still does not defend a workbook
against an agent granted an unconfined shell with an interpreter and a
writable authoritative file; that is the harness\textquotesingle s
responsibility, stated as a scoping assumption, not a guarantee.

\emph{Only value and formula edits; structural edits are unbuilt --- the
key open problem.} The surgical writer edits
\texttt{\textless{}c\textgreater{}} elements in place and exposes only
\texttt{set\_cell} and \texttt{set\_formula}; row/column insert and
delete, sheet add/remove, and range moves --- the structurally
corruption-prone operations, and the ones a shared-formula, merged-cell,
and reference-rewriting model makes hardest to keep surgical --- are
named as the paper\textquotesingle s central open problem, not
implemented, so the fidelity, agent-case-study, and non-bypassability
results are all scoped to value and formula edits.

\emph{The census drops phonetic runs.} The structural census and the
engine\textquotesingle s import path do not preserve furigana
\texttt{rPh} runs, so \texttt{PHONETIC} evaluates against base text and
carries no oracle signal (LibreOffice does not implement it either).

\subsection{12. Conclusion}\label{12-conclusion}

This paper set out to give an LLM agent a way to \emph{edit} a
spreadsheet that it cannot use to corrupt one, for an artifact class
whose owner is accountable for every cell. The core is a surgical
transactional write with a property that holds by construction and is
enforced on every write: after an edit, every OOXML part without a
changed cell is byte-identical to the input, because the writer copies
those parts verbatim and re-serializes only the sheet parts an operation
touches --- and because each apply byte-diffs the un-edited parts and
aborts on any change before it commits. Four things were learned in
building and testing it. First, the property is real and it is the
escape from the fidelity gap of §2: the only way to keep byte-identity
as an integrity signal is to stop rewriting the whole file, which no
whole-file substrate does. Second, the property survives contact with a
real agent: in an interventional case study of two tasks, a status-quo
agent reproduced the \#22044 corruption live in one observed trajectory
and reported success on a non-reloadable file, while an agent confined
to xlq made the identical edit with charts, pivot, and VBA
byte-identical and left a receipt. Third, the boundary\textquotesingle s
honesty is machine-checkable and falsifiable: the runtime declares per
file what it cannot evaluate, the write is refused when its prediction
is unreliable, and an independent financial cross-check corroborated two
of the engine\textquotesingle s verdicts and overturned a third ---
which we report rather than hide. Fourth, an evaluation can measure
fidelity; AXLE-bench adds the axis the field\textquotesingle s
benchmarks omit. We claim no new cryptography, transaction model, or
differential-testing technique --- receipts, transactional rollback, and
cross-engine differential testing are established --- no formal proof,
and no result beyond demonstration scale. The contribution is to fold
the artifact\textquotesingle s format fidelity into an agent write
boundary, surgically and with an enforced-and-checked fidelity property,
and to demonstrate under a real agent --- at the scale of a case study
--- that an agent confined to it cannot commit the corruption that
shipped in production, for the in-place value and formula edits it
covers; structural edits remain the key open problem. The earlier
version of this work could observe that corruption; this one prevents
it, within the scope it honestly claims.

\subsection{References}\label{references}

\emph{Runtime enforcement for LLM agents}

\begin{itemize}
\tightlist
\item
  \textbf{{[}Progent 2025{]}} Progent: Programmable Privilege Control
  for LLM Agents. 2025.
\item
  \textbf{{[}AgentSpec 2025{]}} AgentSpec: Runtime Enforcement of
  LLM-Agent Safety via a Domain-Specific Rule Language. 2025.
\item
  \textbf{{[}MiniScope 2025{]}} MiniScope: Least-Privilege OAuth Scope
  Derivation for Agents via Integer Programming. 2025.
\item
  \textbf{{[}IsolateGPT 2024{]}} IsolateGPT: An Execution Isolation
  Architecture for LLM-Based Agentic Systems. 2024.
\item
  \textbf{{[}Fault-Tolerant Sandboxing 2025{]}} Fault-Tolerant
  Sandboxing for LLM Agents: ACID Snapshots with 100\% Filesystem
  Rollback. 2025.
\item
  \textbf{{[}Cordon 2025{]}} Cordon: Semantic Transactions for
  Multi-Step Agent Actions. 2025.
\item
  \textbf{{[}SAFEFLOW 2025{]}} SAFEFLOW: Compensating-Transaction Safety
  for Agent Workflows. 2025.
\end{itemize}

\emph{Receipts and provenance}

\begin{itemize}
\tightlist
\item
  \textbf{{[}Notarized Agents 2026{]}} Notarized Agents:
  Receiver-Attested Cryptographic Receipts for Agent Actions. 2026.
\item
  \textbf{{[}Nitro 2025{]}} Forward-Secure MAC Logging for
  Tamper-Evident Audit Trails. ACM CCS, 2025.
\item
  \textbf{{[}Certificate Transparency 2022{]}} SoK: Certificate
  Transparency and Merkle-Proof Audit Logs. IEEE S\&P, 2022.
\item
  \textbf{{[}Moreau et al. 2013{]}} L. Moreau et al. The PROV Data Model
  (W3C PROV) and PROV-AGENT. W3C Recommendation, 2013.
\end{itemize}

\emph{Differential testing}

\begin{itemize}
\tightlist
\item
  \textbf{{[}Yang et al. 2011{]}} X. Yang, Y. Chen, E. Eide, J. Regehr.
  Finding and Understanding Bugs in C Compilers (CSmith). PLDI, 2011.
\item
  \textbf{{[}Le et al. 2014{]}} V. Le, M. Afshari, Z. Su. Compiler
  Validation via Equivalence Modulo Inputs (EMI). PLDI, 2014.
\item
  \textbf{{[}Rigger and Su 2020{]}} M. Rigger, Z. Su. Testing Database
  Engines via Pivoted Query Synthesis (PQS/SQLancer) and NoREC. OSDI /
  ESEC-FSE, 2020.
\item
  \textbf{{[}Park et al. 2021{]}} J. Park et al. JEST: N+1-Version
  Differential Testing of JavaScript Engines with a Mechanized
  Specification. ICSE, 2021.
\item
  \textbf{{[}SmartOracle 2026{]}} SmartOracle: LLM-Mediated Triage of
  JavaScript-Engine Disagreements Against the ECMAScript Specification.
  2026.
\item
  \textbf{{[}Liu et al. 2024{]}} Liu et al. Specification-Based
  Differential Testing of SQL Engines with a Three-Way Verdict. 2024.
\item
  \textbf{{[}Almiron et al. 2010{]}} M. Almiron et al. On the Numerical
  Accuracy of Spreadsheets. Journal of Statistical Software, 2010.
\item
  \textbf{{[}McCullough 2008{]}} B. D. McCullough. Assessing the
  Reliability of Statistical Software in Spreadsheets. 2008.
\end{itemize}

\emph{Spreadsheet research and LLM spreadsheet agents}

\begin{itemize}
\tightlist
\item
  \textbf{{[}Panko 2008{]}} R. R. Panko. What We Know About Spreadsheet
  Errors. 2008.
\item
  \textbf{{[}Hermans and Murphy-Hill 2015{]}} F. Hermans, E.
  Murphy-Hill. Enron\textquotesingle s Spreadsheets and Related Emails.
  ICSE, 2015.
\item
  \textbf{{[}Fisher and Rothermel 2005{]}} M. Fisher, G. Rothermel. The
  EUSES Spreadsheet Corpus. 2005.
\item
  \textbf{{[}Barowy et al. 2018{]}} D. W. Barowy, E. D. Berger, B. Zorn.
  ExceLint: Automatically Finding Spreadsheet Formula Errors. OOPSLA,
  2018.
\item
  \textbf{{[}Li et al. 2023{]}} H. Li et al. SheetCopilot. NeurIPS,
  2023.
\item
  \textbf{{[}SheetAgent 2025{]}} SheetAgent: A Generalist Agent for
  Spreadsheet Reasoning and Manipulation. WWW, 2025.
\item
  \textbf{{[}SpreadsheetBench 2024{]}} SpreadsheetBench: Towards
  Challenging Real-World Spreadsheet Manipulation. NeurIPS D\&B, 2024.
\item
  \textbf{{[}SpreadsheetBench 2026{]}} SpreadsheetBench 2: Successor
  Benchmark. 2026.
\item
  \textbf{{[}Finch 2025{]}} Finch: Evaluating LLM Agents on Real-World
  Finance Spreadsheet Workflows. 2025.
\item
  \textbf{{[}OS-Harm 2025{]}} OS-Harm: Measuring Harmful Side Effects of
  Computer-Use Agents. NeurIPS, 2025.
\item
  \textbf{{[}Wang et al. 2021{]}} Z. Wang et al. TUTA: Tree-Based
  Transformers for Generally Structured Table Pre-Training. KDD, 2021.
\item
  \textbf{{[}Dong et al. 2024{]}} H. Dong et al. SpreadsheetLLM:
  Encoding Spreadsheets for Large Language Models. 2024.
\item
  \textbf{{[}Pista 2026{]}} Pista: Step-Level Human Oversight of
  Spreadsheet Agents. 2026.
\end{itemize}

\emph{Standards and artifacts}

\begin{itemize}
\tightlist
\item
  \textbf{{[}claude-code\#22044{]}} Anthropic. claude-code issue
  \#22044: xlsx agent skill corrupts financial models (charts, pivots,
  VBA) via openpyxl. GitHub, 2025.
\item
  \textbf{ECMA-376 / ISO-IEC 29500.} Office Open XML File Formats.
  (Standardizes the OOXML container, not formula numerical semantics.)
\item
  \textbf{OpenFormula.} OpenDocument Formula Specification (ODF). OASIS.
  (Normative for ODF, not Excel\textquotesingle s dialect.)
\end{itemize}

\end{document}
