% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
\documentclass[
  11pt,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
% fallback for those not using the hyperref driver hyperxmp:
\makeatletter
\@ifundefined{xmpquote}{\newcommand{\xmpquote}[1]{#1}}{}
\makeatother
\hypersetup{
  pdftitle={Certify-or-Refuse: A Machine-Checked Soundness Boundary for Untrusted Agent Edits to Opaque-Semantics Artifacts},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Certify-or-Refuse: A Machine-Checked Soundness Boundary for
Untrusted Agent Edits to Opaque-Semantics Artifacts}
\author{}
\date{2026-07-09}

\begin{document}
\maketitle

\section{Abstract}\label{abstract}

An LLM agent that edits a spreadsheet, a database schema, or a dbt
project produces an artifact that \emph{opens fine} but may be silently
wrong: a structural edit that fails to propagate references corrupts
computed values with no visible symptom, and neither the agent nor a
human reviewer can see it without the defining engine. We ask which such
edits can be certified \textbf{engine-free} --- offline, against the
artifact's own structure, trusting neither the editor nor a
recomputation --- and we answer by machine-checking \textbf{both sides
of the boundary} (Lean 4, no \texttt{sorry}, axioms \texttt{propext} and
\texttt{Quot.sound} only), with the remaining middle ground stated open.
On the certifiable side: the premise of our invariance theorem --- the
edit's reference-dependency graph is a function-preserving relabeling of
the original's --- is \emph{decidable from syntax}, and we ship an
executable decision procedure \texttt{check} with a machine-checked
soundness theorem: if \texttt{check} accepts, every computed value
transports across the edit under \textbf{every possible engine} (the
engine is universally quantified, never run), carrying the artifact's
embedded cached values as a self-oracle. On the uncertifiable side: an
impossibility theorem --- an edit introducing function semantics
unwitnessed in the original admits two engines that are pointwise
indistinguishable through the original artifact yet disagree on the
edited value, so \emph{no} engine-free checker can certify it.
Certify-or- refuse is therefore not a design choice but the only sound
shape, and our checker's accept class is exactly the
\emph{proven-certifiable} relabeling class --- every accept is
certifiable; whether the certifiable class extends further (edits
reusing witnessed semantics) is stated open. A companion locality
theorem factors mixed edits into a certified structural scaffold plus a
bounded value-fill audit cone. The theory is load-bearing in the running
system: the deployed checker agrees with the Lean decision procedure on
a randomized differential battery, a production certifier
(\texttt{xlq\ certify}) covers five structural operations on real
spreadsheets, and the same format-parametric core certifies dbt model
refactors engine-free --- catching dangling references and silent logic
changes that re-materialization confirms, while certifying faithful
renames with no SQL executed. We are deliberate about what is
\emph{proof} and what is \emph{corroboration}: the machine-checked
boundary is the result; the empirical harness (147/147 corrupted foreign
edits refused; a naive edit path silently corrupts 85.5\% of real
workbooks (confirmed-genuine, after our own study corrected two
mislabels) while the certified path corrupts none; every injected
corruption in a diverse-corruptor battery refused) corroborates it, and
we report each measurement with its confound --- including two false
certifications our own adversarial reviews found in earlier versions of
the system, which we closed and report as fixed defects.

\section{1. Introduction}\label{introduction}

The scarce resource in the agentic era is not capability but
\textbf{verifiability}. A capable model will, with high probability,
perform a structural spreadsheet edit correctly; but ``high
probability'' is exactly the wrong guarantee for a financial model, a
payroll sheet, or a regulatory filing, where a single unpropagated
reference silently changes a computed total and ships. The failure is
invisible by construction: the edited file is a valid \texttt{.xlsx}, it
opens, and the wrong numbers sit in cells whose formulas \emph{look}
plausible. You cannot see the corruption without the engine, and asking
the engine is asking the very component whose edit you are trying to
check.

This paper asks a narrower, more durable question than ``can the model
do the edit?'': \textbf{can we certify that a given edit is
value-faithful, offline, against the artifact's own structure, without
trusting the editor and without running the defining engine?} The answer
we prove is yes for the \emph{structural} fragment --- the
coordinate-relabeling edits (insert/delete rows and columns) that are
simultaneously the most common agent operations and the ones whose
damage is most invisible. The certifier accepts a structural edit only
when its reference-dependency graph is the relabeling of the original's,
and refuses otherwise. Because the guarantee is grounded in the artifact
rather than the model, it does not decay as models improve --- it is the
boundary a capable agent should be \emph{required} to pass, not a crutch
for a weak one.

Our claims, in order of strength:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{(The boundary, both sides machine-checked.)} We \emph{bracket}
  engine-free certification of agent edits: function-preserving
  relabelings of the reference-dependency graph are certifiable ---
  witnessed by an executable decision procedure \texttt{check} for that
  premise with a machine-checked soundness theorem quantifying over
  \textbf{all} engines --- while edits introducing function semantics
  unwitnessed in the original are uncertifiable by \emph{any}
  engine-free checker (impossibility: indistinguishable engines
  disagree), so \emph{certify-or-refuse is the only sound shape} (§3,
  Lean 4, no \texttt{sorry}). The middle ground (edits reusing witnessed
  semantics at new positions) is stated open.
\item
  \textbf{(Composition.)} A locality theorem factors a mixed edit into a
  certified structural scaffold plus value fills whose effect is
  provably contained in their downstream cone --- collapsing the audit
  surface from the whole artifact to a bounded set (§3, measured in §5).
\item
  \textbf{(Mechanism, theory-linked.)} A certify-or-refuse router whose
  deployed checker implements \texttt{check} and agrees with the Lean
  decision procedure on a randomized differential battery; a production
  certifier (\texttt{xlq\ certify}, five structural ops) in
  translation-validation mode; a hardened, engine-corroborated trusted
  parse; and fail-closed boundaries for everything outside the verified
  surface (§4).
\item
  \textbf{(Generality + corroboration, reported with confounds.)} The
  same format-parametric core certifies dbt model refactors engine-free
  (§5.6); an engine-free foreign-edit battery, an independent-oracle
  A/B, and a diverse-corruptor confusion matrix support the theory on
  real artifacts, each reported with its limitation, not as independent
  proof (§5).
\end{enumerate}

A recurring, honest theme: three rounds of adversarial review of our own
artifacts found real defects --- silent-corruption bugs in the trusted
tokenizer, a circular experiment, and an unimplemented soundness
boundary --- each of which we fixed and report as a fixed defect,
because the discipline of finding them is part of the contribution (§6).

\section{2. The problem: opaque-semantics artifacts and invisible
structural
damage}\label{the-problem-opaque-semantics-artifacts-and-invisible-structural-damage}

A spreadsheet is a program whose semantics live in an engine (Excel,
LibreOffice Calc, IronCalc) that most tools do not reimplement. An edit
tool that manipulates the file's bytes therefore edits a program it
cannot evaluate. The dominant failure mode is the \textbf{unpropagated
reference}: insert a row above a formula's input and, unless every
reference below the insertion point is shifted, the formula now reads
the wrong cell. The popular \texttt{openpyxl} library's
\texttt{insert\_rows} does exactly this --- it moves cell contents but
rewrites zero references --- so on any workbook with a below-insertion
reference, it produces a file that opens cleanly and computes wrong
(§5.2).

The artifact carries its own partial ground truth: Excel writes a cached
value \texttt{\textless{}v\textgreater{}} next to each formula. This
\emph{self-oracle} is what a certifier can check against --- but only
for cells the artifact chose to cache, and only if the checker can
recompute, which returns us to the engine. Our theorem removes the
recompute: for the structural fragment, value-faithfulness follows from
structure alone.

\section{3. The formal core (the
contribution)}\label{the-formal-core-the-contribution}

We model a computation as \texttt{C\ =\ (fn,\ deps)}: \texttt{fn\ n}
maps a node's ordered dependency \emph{values} to its value, and
\texttt{deps\ n} is its ordered dependency \emph{nodes}. This is exactly
what a formula carries --- what it reads and how it combines what it
reads --- with \texttt{fn} left an arbitrary (opaque) function.
Evaluation is fuel-bounded: \texttt{eval\ C\ 0\ \_\ =\ default} and
\texttt{eval\ C\ (k+1)\ n\ =\ fn\ n\ (map\ (eval\ C\ k)\ (deps\ n))}.
The theorem holds at every fuel level, hence for the true evaluation of
any acyclic computation.

\textbf{Theorem 1 (exact structural certification).} Let \texttt{σ} map
every node of \texttt{C} to a node of \texttt{C\textquotesingle{}} with
the same function and dependency list relabeled by \texttt{σ} (a
function- and edge-preserving graph isomorphism):
\texttt{C\textquotesingle{}.fn\ (σ\ n)\ =\ C.fn\ n} and
\texttt{C\textquotesingle{}.deps\ (σ\ n)\ =\ map\ σ\ (C.deps\ n)}. Then
\texttt{eval\ C\textquotesingle{}\ k\ (σ\ n)\ =\ eval\ C\ k\ n} for all
\texttt{k,\ n}.

\emph{Consequence.} If the original's evaluation is the artifact's
embedded ground truth \texttt{O} (\texttt{eval\ C\ k\ =\ O}, the
self-oracle), then a structurally-faithful edit reproduces that ground
truth at the relabeled positions,
\texttt{eval\ C\textquotesingle{}\ k\ (σ\ n)\ =\ O\ n}, established
without the defining engine and without recomputing \texttt{O}. This is
the exact tier of certify-or-refuse.

\textbf{Theorem 2 (locality / audit-surface bound).} Define
\texttt{agree\_upto\ C\ C\textquotesingle{}\ k\ n}: the two computations
have identical \texttt{fn} and \texttt{deps} at every node within
\texttt{k} dependency-steps of \texttt{n}. Then
\texttt{eval\ C\ k\ n\ =\ eval\ C\textquotesingle{}\ k\ n} --- a node's
value depends only on its dependency cone. Hence a value edit changes
nothing outside its downstream cone: a mixed edit factors into a
certified structural scaffold plus value fills whose effect is provably
contained, collapsing the audit surface from ``did this corrupt anything
anywhere?'' to ``are these N cells right?''.

Both theorems are machine-checked in Lean 4, self-contained (no
Mathlib), with \texttt{\#print\ axioms} reporting only
\texttt{{[}propext,\ Quot.sound{]}} and no \texttt{sorry}
(\texttt{formal/SelfOracle.lean}). The reference-shift algebra's
arithmetic laws (insert∘delete = identity, monotonicity, the six-case
delete clamp against the set-theoretic truth) are separately proved for
all inputs by the Z3 SMT solver (\texttt{formal/shift\_laws.py}).

\textbf{Theorem 3 (graph preservation --- the shift discharges the
hypothesis, for single cells and range endpoints).} Theorems 1 and 2
\emph{assume} the edited graph is the \texttt{σ}-relabeling of the
original's; the tool does not assume it, it produces it by shifting
references. We model a formula as a list of tokens --- each an opaque
literal (number, string, function name), a single cell reference, or a
range carrying its two endpoints --- and machine-check that the
reference-shift produces exactly the relabeled graph:
\texttt{refs\ (shiftF\ σ\ f)\ =\ (refs\ f).map\ σ}, with literals
provably untouched (\texttt{formal/RefShift.lean}). This is
\emph{verbatim} the \texttt{hdeps} premise of Theorem 1, so for these
token shapes the graph isomorphism is discharged \textbf{constructively
by the tool's operation}, not assumed. We also machine-check
\texttt{delete∘insert\ =\ id} on the cell maps. All \texttt{sorry}-free,
axioms \texttt{{[}propext,\ Quot.sound{]}}.

\textbf{What Theorem 3 does and does not give.} It is a fusion law
parametric in an \emph{arbitrary total} map
\texttt{σ\ :\ Cell\ →\ Cell}: it proves the shift \emph{transports} the
reference graph by whatever \texttt{σ} it is given --- it does
\textbf{not} verify that \texttt{σ} is Excel's shift arithmetic (that is
the separate Z3-proved algebra), nor does the total-map model express
the asymmetric six-case delete \emph{clamp} (head-in-band → k,
tail-in-band → k−1), the \texttt{\#REF!} a fully-consumed reference
produces, or the true multi-cell interior of a range (we model a range
by its endpoints). So the honest statement is: \textbf{the
token→graph→value chain is machine-checked for single-cell references
and range-endpoint pairs under a value-preserving relabeling}; the
explicitly \emph{trusted} surface is the byte→token parse (§4, §5.1),
the correctness of \texttt{σ} itself, the asymmetric clamp /
\texttt{\#REF!} algebra, range-interior dependencies, and
sheet-qualified/3D/table/external references. This is a real narrowing
of the trusted base --- not its elimination. Fidelity of value edits
remains out of the exact tier by construction.

\textbf{Theorem 4 (the premise is decidable, and the decider is proven
sound --- \texttt{formal/Checker.lean}).} Theorems 1--3 would still be
decorative if the running system never checked their hypothesis. We
therefore state the hypothesis of Theorem 1 as an \textbf{executable
decision procedure} over syntax alone: \texttt{check\ S₀\ S₁\ σ}
verifies that every checked node's opaque function \emph{skeleton} is
preserved (\texttt{S₁.skel\ (σ\ n)\ =\ S₀.skel\ n}), that its dependency
list is the \texttt{σ}-image
(\texttt{S₁.deps\ (σ\ n)\ =\ map\ σ\ (S₀.deps\ n)}), and that the
checked domain is closed under dependencies. The soundness theorem,
\texttt{check\_sound}, then proves: if \texttt{check} returns
\texttt{true}, every checked node's value transports across the edit
\textbf{for every interpretation \texttt{I} of the skeletons} --- the
engine is universally quantified and never run --- and
\texttt{check\_transports\_oracle} carries the artifact's embedded
cached values to the edited positions. The procedure \emph{executes}
(\texttt{\#eval} in the development runs it: a faithful relabeling
returns \texttt{true}; an argument-order botch and an operator botch
return \texttt{false}). The consequence for the system is architectural:
a certifier implementing \texttt{check} accepts \textbf{any producer's
faithful edit} --- regardless of byte-level differences, cache handling,
or which tool made it --- and its soundness is this theorem, not
equality to a reference transform.

\textbf{Theorem 5 (impossibility --- the other side of the boundary,
\texttt{formal/Impossibility.lean}).} An edit that introduces a function
skeleton \emph{not witnessed anywhere in the original artifact} cannot
be value-certified by any engine-free checker. The proof is
indistinguishability: overriding an engine at an unwitnessed skeleton
changes no evaluation of the original artifact
(\texttt{eval\_override\_fresh} --- the two worlds agree at every node
and every fuel, hence realize every cached observation identically), yet
the edited node's value can be made to differ from \emph{any} value a
checker might commit to (\texttt{fresh\_skeleton\_uncertifiable}), and
two such worlds disagree with each other
(\texttt{two\_worlds\_disagree}) --- and the quantification over
checkers is itself a machine-checked object:
\texttt{no\_engine\_free\_predictor} models a checker's value commitment
as \emph{any} function of the two syntactic artifacts and proves every
such predictor is refuted by an engine consistent with the original.
Sound checkers \textbf{must refuse}. One scope note, stated plainly: the
impossibility is for \emph{semantics-agnostic} checkers --- a checker
carrying a trusted specification of particular function semantics (say,
documented Excel built-ins) has a partial engine and escapes the
hypothesis; that is an oracle assumption, exactly what ``engine-free''
excludes. Within that scope, certify-or-refuse is the only sound shape
--- proven, not a design preference.

\textbf{The boundary, stated.} Together, Theorems 4 and 5 bracket
engine-free certification: edits whose reference graph is a
skeleton-preserving relabeling are certifiable, with an executable
proven-sound decider (Theorem 4); edits introducing unwitnessed
semantics are uncertifiable by any engine-free checker (Theorem 5). The
honest middle ground --- edits that \emph{reuse} witnessed skeletons at
new positions, such as copy-pasting an existing formula shape onto new
inputs --- is not settled by either theorem and is stated as open: its
values are forced by the original's syntax only when the reused
skeleton's dependencies carry observed values, a case our current
checker routes to refusal.

\section{4. The certify-or-refuse router and its trusted
base}\label{the-certify-or-refuse-router-and-its-trusted-base}

\textbf{The router.} Given an original artifact and an untrusted foreign
edit plus a declared structural op, the router computes the op's node
relabeling \texttt{σ}, and CERTIFIES iff the foreign edit's
reference-dependency graph is exactly the \texttt{σ}-image of the
original's; otherwise it REFUSES. Two properties make this sound as a
check of \emph{untrusted} work: under-declaring a change is caught (an
unaccounted graph difference → refuse), and a wrong \texttt{σ} yields a
mismatch → refuse. The router is fail-closed: any condition it cannot
certify becomes a refusal.

\textbf{Two certifier modes, honestly labeled.} The system ships two
implementations of the router, and we are precise about which carries
which guarantee. \emph{(i) Direct-premise mode} implements Theorem 4's
\texttt{check} literally: extract both artifacts into
\texttt{(skeleton,\ deps,\ oracle)} triples and decide the premise ---
skeletons preserved, dependency lists the \texttt{σ}-image, containment
(no unaccounted nodes), oracle values preserved outside any declared
fill cone. Its accept class is exactly the certifiable class of §3 ---
any tool's faithful edit, regardless of bytes --- and its soundness is
\texttt{check\_sound}. The deployed implementation agrees with the Lean
decision procedure (run via \texttt{\#eval}) on a randomized
differential battery of faithful edits and four botch classes ---
skeleton change, dependency reorder, retarget, dropped dependency --- at
30/30 (\texttt{formal/differential\_check.py}); we state plainly that
this differential link is evidence of implementation fidelity, not a
proof of it. \emph{(ii) Translation-validation mode} is the production
Rust path (\texttt{xlq\ certify}): apply the tool's own proven
structural transform to the original and diff the foreign edit against
it, certifying iff the only differences are stripped caches and number
formats. It is stricter than mode (i) (byte-for-formula identity to the
canonical transform), engine-free, and multi-sheet-safe; its marginal
value over \emph{doing} the edit is trust-topology --- the untrusted
producer's artifact ships, gated fail-closed, with the tool off the
write path --- not a verifier-cheaper-than-prover asymmetry (there is
none in this mode; it recomputes the transform). Mode (i) is where the
theory is load-bearing; mode (ii) is the hardened production gate.

\textbf{The trusted base: the reference-shift tokenizer.} Every
certificate is only as sound as the predicate deciding \emph{which
tokens in a formula are cell references}. We initially used syntactic
proxies (a 1--3-letter column, an unbounded row) and adversarial review
found these were wrong in three ways that caused \emph{silent
corruption}: a function name whose prefix looks like a cell
(\texttt{BIN2DEC} → \texttt{BIN3DEC} under a row insert), a function
name ending in digits before a paren (\texttt{LOG10} → \texttt{LOG11}),
and out-of-grid tokens (\texttt{XFE9}, \texttt{A2000000}) whose column
or row exceeds the sheet limits. We replaced the proxies with the exact
\textbf{grid-validity} predicate --- a token is a cell reference iff its
column is in \texttt{A..XFD} (1..16384) and its row is in
\texttt{1..1048576}, boundary-gated so an identifier or function call is
never mistaken for a reference. This one rule subsumes all four bug
classes.

We validate the tokenizer at two levels. First, \textbf{differential
cross-checking} against an independent A1 shifter over 19 digit-bearing
Excel functions, out-of-grid tokens, \texttt{\$}-absolutes, and ranges,
across insert-rows and delete-rows: 175 (formula, op) pairs, 0
disagreements (\texttt{tokenizer\_fuzz.py}). This establishes
\emph{mechanization consistency} between the Rust scanner and the
reference --- but the reference is a second implementation of \emph{our
own} grid-validity spec, so it cannot establish conformance to a
spreadsheet engine's semantics. Second, therefore, \textbf{differential
value-preservation against an independent engine}: a blank-row insert is
value-preserving under a correct reference shift, so recomputing the
same file with LibreOffice \emph{before and after} the edit must leave
every formula's value unchanged (\texttt{tokenizer\_conformance.py},
seeded property-based generation over evaluable formulas ---
digit-bearing function names, single/mixed/absolute refs, ranges,
strictly-distinct cell values so a mis-shift onto another cell changes
the value). We run this across \textbf{all four structural ops}
(insert/delete × rows/columns) and against \textbf{two independent
engines} --- LibreOffice \emph{and} the pure-Python \texttt{formulas}
engine, neither of which is xlq's IronCalc
(\texttt{conformance\_v2.py}): \textbf{465 formulas checked per engine,
0 value divergences, with the two engines agreeing exactly.} (The
earlier single-op/one-engine run, \texttt{tokenizer\_conformance.py},
gave 264 formulas / 0 divergences.) We are precise about what this does
and does not establish. Value-preservation is \emph{necessary but not
sufficient} for reference-graph correctness: a mis-shift of a reference
that the formula's value does not depend on (a zero-weighted term, a
non-max argument of \texttt{MAX}) would pass undetected, so the count
over-counts references actually \emph{witnessed} by the value. Two
independent engines agreeing rules out a single-engine artifact, but
both are open-source engines, not Excel; and
sheet-qualified/3D/table/R1C1 constructs are outside the generator. The
honest label is \emph{differential value-preservation against
LibreOffice --- a necessary-not-sufficient corroboration of the trusted
parse, complementing the same-spec fuzzer} --- not full conformance to
an engine's reference semantics.

\textbf{The one undecidable case, made a fail-closed boundary.} A
defined name spelled like a grid-valid cell (\texttt{FY2021} = column
FY, row 2021; \texttt{Q1}; \texttt{Tax2020}) is indistinguishable from a
reference \emph{in the formula text}, so the tokenizer would shift its
uses and the edited file would still equal the tool's own (wrong)
transform --- the single place \texttt{certified\ ⇒\ correct} could be
false on a real workbook. It is, however, decidable from the workbook's
defined-names table, which the certifier already holds. We close it
fail-closed: a structural edit on a workbook containing a defined name
that collides with a cell-reference spelling emits a
\texttt{defined\_name\_ref\_collision} residual, and the certifier
refuses. A workbook defining \texttt{FY2021} and using it in a formula
routes to REFUSE; a benign name (\texttt{TaxRate}) proceeds. This
converts the last hidden hole into a demonstrated boundary and scopes
the verified surface explicitly: \textbf{single sheet, in-grid
coordinates, row/column shifts, no name collision.}

\section{5. Corroboration on real workbooks (reported with
confounds)}\label{corroboration-on-real-workbooks-reported-with-confounds}

None of the following is the soundness argument --- that is §3. Each
supports the theorem on real files and each carries a limitation we
state.

\subsection{5.1 Engine-free certification of untrusted foreign
edits}\label{engine-free-certification-of-untrusted-foreign-edits}

Running the router (format-parametric core) on \texttt{openpyxl}'s edits
of 172 real formula-bearing workbooks, engine-free, cross-checked
against an independent LibreOffice oracle: \textbf{0 false
certifications and 147/147 corrupted foreign edits refused}
(\texttt{foreign\_certify}). The load-bearing caveat, always reported
alongside: this soundness is bought with heavy refusal --- the
fail-closed A1 proxy certifies only about \textbf{1 in 23}
\emph{faithful} foreign edits, because it refuses every reference form
it cannot fully model (whole-row, whole-column, cross-sheet, table,
defined-name). Useful soundness --- a high certify rate on faithful
edits --- requires the complete parser of §4, not the proxy. The value
is the boundary, not the throughput.

\subsection{5.2 Independent-oracle A/B: the corruption the boundary
prevents}\label{independent-oracle-ab-the-corruption-the-boundary-prevents}

On 172 real workbooks, insert-row@2, with LibreOffice (independent of
both \texttt{openpyxl} and the tool's engine) recomputing each edit
against Excel's cache: the naive \texttt{openpyxl} edit path silently
corrupts 149/172 workbooks \emph{as originally measured} --- of which
\textbf{147/172 = 85.5\% is confirmed-genuine reference corruption}
after our own coincidence-bound study (§5.8) surfaced two label
mislabels (files whose formulas carry \emph{zero references of any
kind}, where the flagged divergence is LibreOffice recomputing
\texttt{ACCRINT} differently from Excel's cache, not corruption; three
further files carry only non-A1 reference classes where corruption is
plausible but not attributable by this oracle). The tool's structural
edit is engine-confirmed faithful on 150/172 and explicitly refuses the
remaining 22 --- \textbf{0 silent corruptions} (\texttt{agent\_ab}). We
extend this across \textbf{all four structural ops and both independent
engines} on generated distinct-value workbooks
(\texttt{agent\_ab\_v2.py}): the naive path silently corrupts
insert-rows 100\% / delete-rows 98.8\% / insert-cols 94.3\% /
delete-cols 80.2\% (\textbf{93.8\% overall}), the tool \textbf{0\% on
every op}, and the \texttt{formulas} engine agrees with LibreOffice
\emph{exactly} --- so the protection is not a LibreOffice artifact and
holds across the op space, not just insert-row@2. We flag the confound
that remains: the openpyxl figure is a corpus/tool property (openpyxl
shifts no references), not an agent-error distribution, and a competent
reference-shifting engine also gets these ops right; the tool's
differentiator is auditability and explicit refusal, which the A/B does
not isolate. We found and fixed an oracle false-positive during this
study --- \texttt{ROW()}/\texttt{COLUMN()} are position-dependent and
legitimately change on a row insert --- which is why the oracle excludes
position-dependent functions.

On the \textbf{real} corpus across all four ops we check shift
correctness \emph{deterministically} rather than by recompute, because
real-corpus value-preservation is confounded --- LibreOffice recomputes
exotic financial/date functions (ACCRINT, CUMPRINC, DB, DAYS360, TIME)
inconsistently with the Excel cache, flagging as ``corrupt'' files whose
shifted formulas we verified correct by hand. So we compare xlq's output
formulas to the two-engine-validated reference shifter over
\textasciitilde6,000 real formula cells
(\texttt{shift\_correctness\_real.py}): \textbf{xlq is 100\% correct on
every op} (insert-rows 1651 cells, delete-rows 1608, insert-cols 1648,
delete-cols 1076; 0 mismatches), while the naive path leaves 17--72\% of
shift-requiring cells wrong. Constructs the simple checker cannot
independently verify (whole-column/row, cross-sheet, tables,
range-with-function endpoint) are skipped, not guessed; every early
``mismatch'' was a checker bug (comparing a deleted-band cell to the
wrong output cell; the reference shifter not handling \texttt{F:F},
which xlq correctly shifts to \texttt{G:G}), and xlq had zero real
errors.

\subsection{5.2b A fifth op --- move-rows --- shows σ is not special to
insert/delete}\label{b-a-fifth-op-move-rows-shows-ux3c3-is-not-special-to-insertdelete}

Theorem 1 holds for \emph{any} function-and-dependency-preserving
isomorphism \texttt{σ}, so the certifiable class is not limited to the
monotonic insert/delete shift. We add \textbf{move-rows} (relocate a
contiguous block of rows), whose \texttt{σ} is a \emph{permutation} ---
proved a bijection in Z3 (injective + surjective on
\texttt{{[}1,\ maxrow{]}}, both directions) --- and whose physical edit
reorders rows (a buffered, non-streaming rewrite). A range whose
endpoints reorder under the permutation cannot be a shifted rectangle,
so it fail-closes as a \texttt{move\_straddles\_range} residual. On the
same real corpus, deterministic move-rows shift correctness is
\textbf{100\% on 1,538 real formula cells} (25/60 workbooks fail-closed
as residual/straddle), and the certify-or-refuse contract holds:
\texttt{certify(orig,\ correct\ move)} CERTIFIES while a botched move
(one reference reverted) is REFUSED. This is a concrete demonstration
that the exact tier extends to a genuinely non-monotonic structural edit
by supplying the right \texttt{σ}, with no change to the theorem.

\subsection{5.3 Independent-oracle confusion matrix for the production
certifier}\label{independent-oracle-confusion-matrix-for-the-production-certifier}

We adjudicate \texttt{xlq\ certify}'s verdicts \emph{independently of
the tool} over \textbf{diverse} corruption. An initial matrix used a
single corruptor --- \texttt{openpyxl}'s deterministic no-op-shift ---
and the Excel cache with a reliability gate; its \texttt{FN=0} on 14
edits was a point estimate (rule-of-three ≈ 21\%) confounded with editor
identity. We diversified to three corruptor types --- \texttt{openpyxl}
(no-op), \texttt{unshift\_one} (revert one shifted reference), and
\texttt{wrong\_delta} (over-shift one reference) --- with ground truth
\textbf{by construction} (we inject the corruption, so its label is
independent of any engine) and a LibreOffice \emph{self-consistent}
oracle as cross-check (recompute before/after; no reliability gate). All
45 injected corruptions were refused (each type 15/15) and all 15
faithful tool-produced edits certified (\texttt{cert\_confusion\_v2}).
We are careful about what this does \emph{not} show: because the
certifier accepts iff the edit equals the tool's transform, and all
three corruptors are \emph{defined} as deviations from that transform,
\texttt{FN=0} here is \textbf{by construction}, not a sampled bound ---
we do not attach a confidence interval to a structurally-guaranteed
zero, and \texttt{unshift\_one}/\texttt{wrong\_delta} differ only in the
sign of a perturbation the certifier does not weigh. The one
\emph{reachable} false certification is not in this family at all: the
cell diff compares only sheet cells, so a foreign edit that shifts every
cell formula correctly but leaves a \textbf{non-cell reference} (a
defined name's target, a data-validation or conditional-formatting
formula, a chart series) unshifted is invisible to it. We built exactly
that edit (\texttt{cert\_noncell\_test.py}): with a defined name
\texttt{Rate\ =\ \$A\$10} left unshifted while all cells moved, an
earlier \texttt{xlq\ certify} \textbf{CERTIFIED it with all-zero diffs
--- a genuine false certification.} We closed it: certify now checks
that defined names match the tool's transform (refusing the unshifted
one, certifying the correct one) and \emph{fail-closes} on
data-validation, conditional-formatting, and chart parts it does not
compare --- a stated coverage boundary. Separately, 6 of the
cell-formula corruptions were \textbf{value-preserving} (a reference
error that does not change the current value); the value oracle called
them faithful but certify refused all 6 --- its structural equality is
\emph{stricter} than a value check. The symmetric cost, stated plainly:
certify equally refuses a \emph{faithful but non-identical} rewrite (a
commuted \texttt{A1+B1} → \texttt{B1+A1}), so its accept class is
``byte-for-formula identical to the tool's transform,'' narrower than
``value-faithful.''

\subsection{5.4 The interventional finding: differential testing
hardened the trusted
base}\label{the-interventional-finding-differential-testing-hardened-the-trusted-base}

We ran a live fast-model agent on 20 real workbooks, tasking it to
rewrite formula references for insert-row@2, and compared its output to
the tool's transform. As a guard-soundness experiment this is
\textbf{circular} --- the ``agent correct'' label and the
\texttt{xlq\ certify} verdict are the same predicate applied to a file
grafted onto the tool's own output --- and we do not report it as such.
Its genuine, non-circular yield is \textbf{differential testing that
surfaced real silent-corruption bugs in the tool's own tokenizer}
(\texttt{BIN2DEC}, \texttt{Sales2020}, and on follow-up \texttt{LOG10}
and the out-of-grid class), each now fixed and covered by §4's
cross-check. A validation method that finds real bugs in its own trusted
base is exactly what a certifier's TCB needs; the honest framing is that
the method proved its worth, not that it produced a soundness number.

\subsection{5.5 Composition on mixed edits,
measured}\label{composition-on-mixed-edits-measured}

By Theorem 2 a mixed task factors into a certified structural scaffold
plus value fills contained in their downstream cone. We measure this on
realistic mixed edits (\texttt{composition\_coverage.py}): every
scaffold is certified, and the audit surface collapses to the fill cone
--- 100/84/68/52\% of the artifact certified untouched as the number of
fills grows (mean 76\%) --- so on a realistic edit-distribution study
the certifiable \emph{component} rises from 27\% of tasks fully
certified to \textbf{87\% with a certified scaffold}.

\subsection{5.6 A second real domain: dbt model refactors, certified
engine-free}\label{a-second-real-domain-dbt-model-refactors-certified-engine-free}

The theory is format-parametric --- nothing in §3 mentions spreadsheets
--- and we cash that on a practitioner-relevant second domain:
\textbf{dbt projects}, where models reference each other via
\texttt{\{\{\ ref(…)\ \}\}}, the everyday edit is a rename/refactor, and
the warehouse's materialized tables from the pre-edit state are the
self-oracle. The adapter extracts the same
\texttt{(skeleton,\ deps,\ oracle)} triple the checker consumes:
skeletons are SQL with references abstracted to ordered slots (case and
whitespace folded \emph{outside} string literals ---
\texttt{\textquotesingle{}ABC\textquotesingle{}} is never conflated with
\texttt{\textquotesingle{}abc\textquotesingle{}}; any unmodeled Jinja
marks the node dynamic and out of the exact tier, refused rather than
guessed); the edited artifact is built with \textbf{no SQL executed}.
Four scenarios on a ten-node staging→marts project: a faithful rename
with all references updated is \textbf{CERTIFIED} (100\% of the artifact
untouched, no engine run); a rename leaving one downstream reference
dangling is \textbf{REFUSED}; a rename with a silently changed
aggregation (\texttt{SUM}→\texttt{AVG}) is \textbf{REFUSED}, and
independent re-materialization confirms the values really differ; a
rename plus a \emph{declared} logic change is certified as a scaffold
with the audit surface exactly the declared model plus its downstream
cone (collapse 0.8), and re-materialization confirms outside-cone values
are preserved. Honest scope: model-level granularity (the cone is whole
models, not columns); a mini-dbt subset (single-argument
\texttt{ref}/\texttt{source}; no macros); and fail-closed normalization
--- a comment-only reformat is refused, never wrongly certified.

\subsection{5.7 The interventional study: live agents, guarded vs
unguarded, with
cost}\label{the-interventional-study-live-agents-guarded-vs-unguarded-with-cost}

Finally, the experiment every earlier version of this work lacked:
\textbf{live agents' own errors}, caught by the direct-premise checker,
with completion cost measured. Design, with the independence structure
stated: 21 real-workbook tasks (196 formula cells; task prep excludes
constructs the guard cannot see, so measured cost \emph{understates}
real-world refusal cost --- stated, not hidden); the agent's artifact is
built via \texttt{openpyxl} plus surgical formula splicing with the
reference tool nowhere in it; ground truth is the two-engine-validated
reference shifter (cells outside its grammar are excluded and counted,
never guessed); the guard is the direct-premise checker of §4.i; and two
live agent conditions (a careful and a hasty prompt, same fast model)
make the error distribution the agent's own rather than injected. The
harness was first validated on synthetic perfect/sloppy agents (perfect:
zero corruption in both arms; sloppy: all seven injected corruptions
blocked).

Results: the careful agent erred on 2/21 tasks, the hasty agent on 4/21
(six saves total across the two arms --- which share tasks and model, so
they are not independent runs --- spanning four distinct workbooks; with
n this small we report counts, not rates: 6/6 blocked has a 95\%
binomial lower bound of 0.54). Unguarded, all six erroneous artifacts
ship as silent corruption; four are content errors and two are
protocol-keying errors (the agent's formulas were right but returned
under shifted addresses --- interface-dependent corruption, still
corruption as shipped). Guarded, \textbf{zero ship --- all six are
refused, with zero false certifications on the 137/196 truth-visible
cells and zero unambiguous completion cost}. The symmetric caveats, both
stated: certified artifacts contain truth-blind cells (11 of 19
certified tasks carry cells outside the truth grammar), where corruption
would be invisible to the truth instrument just as hidden saves are ---
the instrument cannot contradict the guard there in either direction;
and the two refusals of ``correct'' hasty work are on truth-partial
tasks whose unverifiable cells may themselves be wrong --- possible
hidden saves, split-reported. Two findings deserve emphasis. First, the
live error distribution \emph{naturally} contained the adversarial
classes our earlier synthetic corruptors were criticized for omitting:
partial shifts (\texttt{F.INV(F2,B3,C3)} --- two references shifted, one
missed), a protocol misread (formulas returned under already-shifted
addresses), and dropped cells. Second, one careful-agent error was a
\textbf{semantics-changing edit with correctly shifted references} ---
\texttt{TEXTSPLIT(A26,,"")} rewritten as \texttt{TEXTSPLIT(A27,"")},
silently turning a row-delimiter argument into a column-delimiter ---
and the graph check caught it because the function \emph{skeleton}
changed even though every reference moved correctly: precisely the class
a value-spot-check or a reference-only linter would miss.

\subsection{5.8 The probabilistic tier, quantified: the coincidence
bound}\label{the-probabilistic-tier-quantified-the-coincidence-bound}

Everything outside the certifiable class routes to probabilistic
checking against the self-oracle --- and until now ``probabilistic'' was
unquantified. We derive the \textbf{coincidence bound}: the probability
that a \emph{wrong} edit passes a k-cell value check because the misread
cells coincidentally carry colliding values. The naive independent bound
\texttt{q\^{}k} is badly optimistic: on the real corpus the honest,
dependence-aware mixture bound (within-file value repetition dominates)
is \textbf{30×} higher at k=5 and \textbf{\textasciitilde2×10⁴×} at
k=10, and tracks a Monte-Carlo simulation on real formulas within
\textasciitilde25\% at every k. Measured on 230 real first sheets under
Excel read semantics, the off-by-one-row collision rate --- exactly the
naive edit path's error --- is q̂ = 0.178 pooled (file median 0.125, p90
0.40; spreadsheets are full of repeated values). Detection therefore
needs \textbf{k = 5 checked cells for 99\% and k = 9--10 for 99.9\%}
(the mixture bound is itself still mildly optimistic within-file --- the
MC sits just above it at k = 10) against \emph{systemic} errors ---
double the naive prescription. Against \emph{localized} errors the tier
is coverage-bound, not collision-bound: a single mis-routed reference
corrupts one cell plus its cone, so a k-of-N sampled check detects it
with probability at most k/N regardless of q. And the hard limit,
engine-verified \textbf{on this corpus} (engine calc-test suites plus
templates --- not in-the-wild business workbooks): \textbf{of 161
workbooks where the naive path's error is genuinely present, 19 (11.8\%;
conservatively 12 = 7.5\%) pass a full k = N value check} ---
LibreOffice recomputes identical values on a \emph{wrong reference
graph} (aggregates, \texttt{MIN}/\texttt{MAX}, \texttt{MODE}-class
functions absorb the misread). The passing files are dominated by test
suites for exotic non-injective functions, over-represented here by
construction; the floor's magnitude plausibly \emph{shrinks} on
arithmetic-dominated business sheets, and we state that direction rather
than claim transfer. This is the quantified case for the exact tier:
value checking, even exhaustive, cannot close the gap that the graph
check closes by construction.

\section{6. On finding our own
defects}\label{on-finding-our-own-defects}

Successive adversarial reviews of these artifacts each landed a real
hole: the tokenizer's syntactic proxies (silent corruption); the
circular live-agent experiment (a tautological ``0 false
certifications''); the unimplemented defined-name \emph{aliasing}
boundary; and --- most consequentially --- a \textbf{reachable false
certification} in the production certifier itself, where a foreign edit
that shifts every cell formula correctly but leaves a defined name's
target unshifted was certified with all-zero diffs, because the cell
diff never compared non-cell references. We built the exploit, confirmed
it, and closed it (defined names now compared against the tool's
transform; data-validation, conditional-formatting, charts, pivots, and
external links fail-closed). A subsequent review then falsified a
\emph{universal} phrasing of that fix by the same pattern --- a foreign
edit that reverts a
\texttt{mergeCell}/\texttt{hyperlink}/\texttt{autoFilter} reference,
which the transform shifts but the cell diff never compared, was
certified --- so we extended the net to those semantic structural
references and reframed the boundary as an explicitly \emph{enumerated
denylist} whose completeness we do not claim to have proven. Most
recently, the coincidence-bound study (§5.8) surfaced label noise in our
own headline A/B: two ``corrupted'' files carry zero references of any
kind, so their flagged divergence is engine disagreement, not corruption
--- we corrected the headline from 86.6\% to 85.5\% confirmed-genuine
and record the mislabel here. We report each as a fixed defect. This is
part of the contribution: a certify-or-refuse claim is only credible if
its authors have tried hardest to break it --- the record of what broke,
and what the fix was, is the evidence that the remaining boundary is
real, and it is why we are conservative about calling the corroboration
anything more than corroboration, and the denylist anything more than a
denylist.

\section{7. Scope and limitations}\label{scope-and-limitations}

The \textbf{verified surface} the certifier certifies (rather than
refuses) is: row/column insert/delete \emph{and row-block move} (§5.2b),
single-sheet, in-grid coordinates, cell formulas over single-cell and
range-endpoint references, with no defined-name/cell collision and (for
move) no move-straddling range. On top of the sheet-cell diff, certify
explicitly compares the reference-bearing constructs the transform
shifts that a cell diff cannot see --- defined-name targets and
mergeCell/hyperlink/autoFilter \texttt{ref}s --- refusing any that
differ from the transform, and fail-closes on data-validation,
conditional-formatting, sparklines, charts, pivot tables, and external
links. Two honesty caveats, both learned from this artifact's own
failures. First, this compare-and-fail-close net is an
\textbf{enumerated denylist}, not a machine-checked allowlist: its
completeness over value-affecting non-cell references is \emph{asserted,
not proven}, so an un-enumerated value-bearing construct would be
silently certified --- the exact failure the defined-name case exhibited
until a reviewer built the exploit. (We keep certify's compare surface a
superset of the transform's \emph{value-bearing} write surface --- every
reference-bearing construct except the deliberately-excluded view-state
\texttt{dimension}/\texttt{selection}/\texttt{pane}/\texttt{brk} ---
which makes the enumeration mechanically checkable against
\texttt{has\_ref\_attr}, but that superset relation is a code invariant,
not a proof.) Second, the guarantee is over \textbf{computed values}:
pure view-state the transform also shifts --- the used-range
\texttt{dimension}, \texttt{selection}, frozen \texttt{pane}, and page
\texttt{brk}s --- is deliberately \emph{not} compared, because it is
non-value-bearing and foreign tools legitimately vary it; a certified
file may differ from the transform there without affecting any value.
Value edits are out of the exact tier by construction. The
\textbf{trusted base} (narrowed by Theorem 3 but not eliminated) is: the
byte→token parse; the correctness of the shift map \texttt{σ} itself
(the Z3-proved arithmetic, not re-proved in the graph layer); the
asymmetric six-case delete clamp and its \texttt{\#REF!} outcomes; the
multi-cell interior of ranges (modeled by endpoints); the constructs
absent from the token type; and the completeness of the non-cell
denylist. The corroboration is bounded accordingly: value-preservation
is necessary-not-sufficient (though now confirmed across all four
structural ops and by two independent engines that agree, ruling out a
single-engine artifact, and still not Excel); the confusion matrix's
\texttt{FN=0} is by-construction over deviations-from-transform, so the
informative test is the non-cell corruptor of §5.3, which found and
closed a real false certification. The exact tier certifies a measured
37.5\% of operations and 27\% of whole tasks on a realistic
edit-distribution study; the majority of real tasks are mixed and need
the probabilistic tier --- but by Theorem 2 a mixed task factors into a
\emph{certified structural scaffold} plus a bounded value-fill cone, and
we measure this (\texttt{composition\_coverage.py}): on mixed edits the
scaffold is always certified and the audit surface collapses to the
value-fill cone (100/84/68/52\% of the artifact certified untouched as
fills grow, mean 76\%), so the certifiable \emph{component} rises from
\textbf{27\% fully certified to 87\% with a certified scaffold}. A
verified byte-level tokenizer, and a model faithful to the
clamp/\texttt{\#REF!} algebra, are the natural next steps.

\section{8. Related work}\label{related-work}

Spreadsheet analysis and smell detection, structural-edit tools that
reshape files without a fidelity property, and record/replay or
unit-test approaches to spreadsheet correctness all differ from this
work in the same way: they establish behavior on tested inputs, whereas
we prove value-fidelity of the structural fragment for all inputs and
all semantics, and gate untrusted edits on that proof.
\textbf{Translation validation} (Pnueli et al.) and verified compilation
are the closest methodological analogues, and our production mode
(§4.ii) is honestly an instance of TV's weakest form --- accept iff the
output equals a proven reference transform. The delta of this work over
TV is twofold. First, the \emph{direct-premise mode} (§4.i) does what TV
does not: it decides the invariance theorem's hypothesis directly on the
untrusted artifact, so its accept class is the entire
\emph{proven-certifiable} relabeling class (any producer's faithful
relabeling, not only outputs identical to a canonical transform; whether
certifiability extends past that class is stated open), and its
soundness quantifies over \textbf{all} engines --- TV assumes the
semantics it validates against, whereas our setting's defining engine is
opaque and absent. Second, TV has no analogue of our impossibility side:
Theorem 5 shows the refusal half of certify-or-refuse is forced,
bracketing the boundary rather than picking a point on it.
Proof-carrying code shares the shape of ``checkable evidence, small
checker,'' and our decision procedure plays the checker's role with the
certificate being the edit descriptor \texttt{σ} itself.

\section{9. Conclusion}\label{conclusion}

Verifiability, not capability, is the durable constraint on agent edits
to opaque-semantics artifacts. We bracketed --- machine-checked on both
sides, middle ground stated open --- what can be certified about such
edits without the engine: the relabeling class is certifiable, witnessed
by an executable decision procedure proven sound under every possible
engine; anything introducing unwitnessed semantics is not, so
certify-or-refuse is the only sound shape. The theory is load-bearing in
the running system --- the deployed checker agrees with the Lean
decider, the production gate covers five structural operations on real
spreadsheets, and the same core certifies dbt refactors engine-free ---
and every measurement is reported with its confound. The result is an
honest, genuinely-verified boundary over an explicitly-scoped surface:
the boundary a capable agent should be \emph{required} to pass, not a
patch for a weak one --- and one that does not decay as agents improve,
because it is grounded in the artifact, not the model.

\end{document}
